---
title: "Figure_2"
output: html_notebook
---
```{r 0-setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '', fig.width = 6, fig.height = 6,
  results = "hide",
  message = FALSE
)
```

# load libraries

```{r 0-library-loads}
require(GenomicFeatures)
require(GenomicAlignments) ##readGAlignments
require(AnnotationDbi)#loadDb
require(data.table)#fast large dataset manipulation
require(readxl)
require(dplyr)


require(ggplot2)
require(RColorBrewer)
require(gridExtra)

library(gplots)
library(RColorBrewer)
library(limma)

# if heya8 still similar to h9, or can be distinguished
# heatmap for samples
library(ComplexHeatmap)
library(circlize)
library(viridis)
gctorture(FALSE)
library(tximport) # 

library(cowplot)
library(ggplotify)# 
source("utility_function.R")

library(bambu)
```


# load general data
```{r}
cat('Setting working directory')
wkdir <- 'wkdir'
general_list <- readRDS("general_list2023-08-29.rds")
samples_wSpikein <- general_list$samples_wSpikein
cellLines <- general_list$cellLines#c('Hct116','HepG2','K562','A549','MCF7',"H9","HEYA8")
protocolCol <-  general_list$protocolCol#adjustcolor(brewer.pal(8,"Dark2")[1:5],0.7)
protocolVec <-  general_list$protocolVec#c("directRNA","directcDNA","cDNA","PacBio","Illumina")
protocolLabel <- general_list$protocolLabel#c("RNA","PCR-free cDNA","cDNA","PacBio","Illumina")
```


# Fig. 2
## a read count
```{r 1-readcount}
salmon_sr <- readRDS(paste0(wkdir,"output/salmon_sr.rds"))

readCountONT <- readRDS("readCount_guppy_6_4_2_May6.rds")
readCountPacBio <- readRDS("readCount_pacbio.rds")
readCountSpikein <- readRDS("readCount_spikein_May5.rds")
readCountSpikeinSR <- readRDS("readCount_spikein.rds")
readCountSpikeinSR <- readCountSpikeinSR[grepl("Illumina", runname)]

# spikein to be added 
readCount <- do.call("rbind",list(readCountONT, readCountPacBio, readCountSpikein, readCountSpikeinSR))
readCount[, runname := gsub(",","",runname)]
# for PacBio remove sirv-1, change sequin MixA V1 to sequin MixAv2
readCount[grepl("PacBio",runname)&grepl("sequin",runname), runname := gsub("v1.0","V2",runname)]
readCount <- readCount[-which(grepl("PacBio",runname)&grepl("SIRV-1",runname))]

samples <- samples_wSpikein[data.table(runname = readCount$runname), on = "runname"]
samples[grepl("allSpikinReadsCombined",runname), `:=`(
    old_runname = runname,
    protocol = unlist(strsplit(gsub("__","_",runname),"_"))[3],
    cellLine = gsub("v1.0","V1",gsub(",","",unlist(strsplit(gsub("__","_",runname),"_"))[2]))), 
    by = runname]
samples[grepl("allSpikinReadsCombined",runname), `:=`(techRep = paste0("run",1:.N)), by = list(cellLine)]
samples[grepl("allSpikinReadsCombined",runname), `:=`(protocol_type = protocol, bioRep = "replicate1")]


readCount <- samples[readCount, on = "runname"]

readCountSR <-unique(salmon_sr[,.(runname, ntotal)])
setnames(readCountSR, c("runname","ntotal"),c("old_runname","total_reads"))
readCountSR <- samples_wSpikein[readCountSR, on = "old_runname"]



samplesRC_combined <- rbindlist(list(readCount, readCountSR), fill = TRUE)
samplesRC_combined[, `:=`(protocol_type = gsub("-SMRTcell","",protocol_type))]
samplesRC_combined[, protocol_type_factor := factor(protocol_type, c('directRNA','directcDNA','cDNA',"PacBio","Illumina"),
                                             protocolVec)]
#samplesRC_combined <- samplesRC_combined[cellLine %in% cellLines]



plotdata <- samplesRC_combined[,.(protocol_type_factor, runname, total_reads, Platform, demultiplexed,cellLine)]
plotdata[is.na(demultiplexed), demultiplexed := FALSE]
## all de-multiplexed are gridion/minion

library(tidyverse)
#library(see)
source("https://raw.githubusercontent.com/datavizpyr/data/master/half_flat_violinplot.R")

# short read are all demultiplexed 
plotdata[Platform == "Illumina", demultiplexed := TRUE]
plotdata[, special_prep:=ifelse(grepl("allSpik",runname),4,ifelse(demultiplexed,3,ifelse(Platform == "PromethION",2,1)))]

# remove spike-in put it in supplementary 
p_sequencing_depth <- ggplot(plotdata[special_prep != 4], aes(x = protocol_type_factor, y = total_reads/1000000))+
    geom_flat_violin( aes(fill = protocol_type_factor, col = protocol_type_factor),
        alpha = 0.5,
                      position=position_nudge(0.25), 
                      adjust = 2)+
    geom_boxplot( aes(fill = protocol_type_factor), 
                 width = 0.05, outlier.alpha = 0,position=position_nudge(0.25))+
     geom_point(aes(col = protocol_type_factor,shape = as.factor(special_prep)),
               position = position_jitter(0.2),  
               alpha = 0.8)+
     scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec,
                       labels = protocolLabel)+
    scale_shape_manual(values = c(1,0,2),labels = c("MinION/GridION","PromethION","Demultiplexed"),
                       name = "Protocol")+
    xlab("Protocols")+
    ylab(bquote('Number of reads '(10^6)))+ #\nPromethION runs excluded
    scale_y_log10(breaks=c(0.001,0.01,0.1,0.5,1, 3, 5, 10, 20, 
                           40,100),
                  labels =c(0.001,0.01,0.1,0.5,1, 3, 5, 10, 20,
                            40,100))+
    theme_classic()
print(p_sequencing_depth)
```


## b read length
```{r 4-read-length, eval = FALSE}
rdfile <- c(list.files(".", pattern = "readLength_", full.names = TRUE, recursive = TRUE),
            list.files(".", pattern = "readLength.*Illumina", full.names = TRUE, recursive = TRUE))
rdfile_name <- gsub("readLength_|\\.rds","",basename(rdfile))
rlData <- do.call("rbind",lapply(seq_along(rdfile_name), function(r){
    print(r)
    rname <- rdfile_name[r]
    rlData <- readRDS(rdfile[r])
    return(unique(rlData[,list(aveLen=mean(aligned_len)), by =  runname]))
}))
saveRDS(rlData, file = paste0(wkdir,"output_guppy6.4.2/readLengthData_Apr27.rds"))

rdfile <- c(list.files(".", pattern = "readLength_", full.names = TRUE, recursive = TRUE),
            list.files("/mnt/projects/SGNExManuscript/output/", pattern = "readLength.*Illumina", full.names = TRUE, recursive = TRUE))
rdfile_name <- gsub("readLength_|\\.rds","",basename(rdfile))

rlData <- do.call("rbind",lapply(seq_along(rdfile_name)[grep("H9_cDNA_replicate4",rdfile_name)], function(r){
    print(r)
    rname <- rdfile_name[r]
    rlData <- readRDS(rdfile[r])
    return(unique(rlData[,list(aveLen=mean(aligned_len)), by =  runname]))
}))
saveRDS(rlData, file = paste0(wkdir,"readLengthData_sample109.rds"))
rlData <- readRDS(paste0(wkdir,"readLengthData_Apr27.rds"))
rlData <- rlData[!grepl("H9_cDNA_replicate4_",runname)]
rlData_sample109 <- readRDS(paste0(wkdir,"readLengthData_sample109.rds"))
rlData <- rbindlist(list(rlData, rlData_sample109))
saveRDS(rlData, file = paste0(wkdir,"readLengthData_May6.rds"))
```




```{r 4-read-length-figure}
rlData <- readRDS(paste0(wkdir,"readLengthData_May6.rds"))
rlData[grepl("PacBio",runname)&grepl("sequin",runname), runname := gsub("v1.0","V2",runname)]
rlData[, old_runname := runname]
temp <- samplesRC_combined[,.(old_runname, runname)]
setnames(temp, "runname", "new_name")
rlData <- temp[rlData, on = "old_runname"]
rlData[!is.na(new_name), runname := new_name]
rlData[, `:=`(old_runname = NULL, new_name = NULL)]
rlData_mat <- samplesRC_combined[rlData, on = "runname"]
rlData_mat[Platform == "Illumina", demultiplexed := TRUE]

rlData_mat[, special_prep:=ifelse(grepl("allSpik",runname),4,ifelse(demultiplexed,3,ifelse(Platform == "PromethION",2,1)))]
rlData_mat[is.na(special_prep), special_prep := 1]
p_mapped_read_length <- ggplot(rlData_mat[special_prep != 4], 
    aes(x = protocol_type_factor, y = aveLen))+ #/1000
       geom_flat_violin( aes(fill = protocol_type_factor, col = protocol_type_factor),
        alpha = 0.5,
                      position=position_nudge(0.25), 
                      adjust = 2)+
    geom_boxplot( aes(fill = protocol_type_factor), 
                 width = 0.05, outlier.alpha = 0,position=position_nudge(0.25))+
      geom_point(aes(col = protocol_type_factor,shape = as.factor(special_prep)),
               position = position_jitter(0.2),  
               alpha = 0.8)+
     scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec,
                       labels = protocolLabel)+
     scale_shape_manual(values = c(1,0,2,3),labels = c("MinION/GridION","PromethION","Demultiplexed","Spikein"),
                       name = "Protocol")+
    xlab("Protocols")+
    scale_y_log10()+
    ylab("Mean read length")+
    theme_classic()
p_mapped_read_length
```



## c 5 to 3 coverage
```{r coverage, eval = FALSE}
cov.dir <- "coverageSaveRDSApr17"
rdsFiles <- dir(cov.dir, pattern = "^covDT",full.names = TRUE)
rdsFiles2 <- dir(cov.dir, pattern = "^GIS")
rdsFiles3 <-  dir(cov.dir, pattern = "^SGNex")
runname <- gsub("covDT|(\\.rds)","",basename(rdsFiles))
txdbEnsembl91 <- loadDb('hg38_sequins_SIRV_ERCCs_longSIRVs-txdb.sqlite')
tx <- transcripts(txdbEnsembl91)
txInfo <- data.table(tx_name = tx$tx_name, strand = as.character(strand(tx)))
# dt <- data.table(runname = c(runname,rdsFiles2))
# dt[, protocol:=strsplit(runname,"\\_")[[1]][3],by = runname]
# dt[,protocol_type:=gsub("RandomPrimer","",gsub("Stranded","",gsub("PromethionD","d",protocol)))]
# dt[, cellLine:=gsub('k562','K562',strsplit(runname, '\\_')[[1]][2]),by = runname]
covData1 <- do.call("rbind",lapply(rdsFiles, function(s){
    print(s)
    covd <- readRDS(s) 
    if(grepl("allSpik",s)){
        setnames(covd, "run_name","runname")
        covd <- covd[!grepl("ENST",tx_name)]
    }
    # covd[, pos_bin_new:=ifelse(is.na(strand), pos_bin, ifelse(strand=="-", 100-pos_bin+1,pos_bin))]
    covdt <- unique(covd[, list(ave_bin_count=mean(rel_bin_count)), by = list(pos_bin,nread,runname)])
    return(covdt)
}))
covData2 <- do.call("rbind", lapply(rdsFiles2, function(k){
    rds <- dir(paste0(cov.dir, "/",k), full.names = TRUE)
    covd <- do.call("rbind",lapply(rds, function(r){
        return(readRDS(r))
    }))
    nread <- sum(unlist(lapply(rds, function(r){
        return(unique(readRDS(r)$nread))
    })))
    covdt <- unique(covd[, list(ave_bin_count=mean(rel_bin_count)), by = list(pos_bin,run_name)])
    covdt[, nread := nread]
    return(covdt)
}))
setnames(covData2, "run_name","old_runname")
covData2 <- samples_wSpikein[,.(old_runname, runname)][covData2, on = "old_runname"]
covData2[, old_runname := NULL]
covData3 <- do.call("rbind", lapply(rdsFiles3, function(k){
    print(k)
    rds <- dir(paste0(cov.dir, "/",k), full.names = TRUE)
    covd <- do.call("rbind",lapply(rds, function(r){
        return(readRDS(r))
    }))
    nread <- sum(unlist(lapply(rds, function(r){
        return(unique(readRDS(r)$nread))
    })))
    covdt <- unique(covd[, list(ave_bin_count=mean(rel_bin_count)), by = list(pos_bin,run_name)])
    covdt[, nread := nread]
    return(covdt)
}))
setnames(covData3, "run_name","runname")
covData <- do.call("rbind",list(covData1, covData2, covData3))
covData[, normCount:=ave_bin_count/max(ave_bin_count), by = runname]
saveRDS(covData, file = paste0(wkdir,"covData.rds"))
```


```{r}
covData <- readRDS(paste0(wkdir,"covData.rds"))
covData[grepl("sequinMixAv1.0_PacBio",runname), runname := gsub("v1.0","V2",runname)]
covData <- samplesRC_combined[, .(protocol_type_factor, cellLine,runname)][covData, on = "runname"]
plotData <- unique(covData[, list(aveNorm=mean(ave_bin_count)), by = list(pos_bin,protocol_type_factor, cellLine)])
plotData[, normCount:=aveNorm/max(aveNorm), by = list(protocol_type_factor,cellLine)]


plotData[, mean_count :=mean(normCount), by = list(protocol_type_factor,pos_bin)]
plotData[, min_count := min(normCount), by = list(protocol_type_factor,pos_bin)]
plotData[, max_count := max(normCount), by = list(protocol_type_factor,pos_bin)]

plotData[, pc := paste0(protocol_type_factor," ",cellLine)]
# p_coverage_5to3_spikein <- ggplot(plotData[grep("sequin|SIRV",cellLine)], aes(x = pos_bin, y = normCount, group = pc))+
#     geom_line(aes(col = protocol_type_factor))+
#     scale_x_continuous(limits = c(0,100),
#                        breaks = c(0,100),
#                      labels = c("5'","3'"))+
#     scale_fill_manual(values = protocolCol,
#                       name = "Protocol",
#                       limits = protocolVec,
#                       labels = protocolLabel)+
#     scale_color_manual(values = protocolCol,
#                        name = "Protocol",
#                        limits = protocolVec[1:5],
#                        labels = protocolLabel[1:5])+
#     ylab("Coverage")+
#     xlab("Transcription direction")+
#     theme_classic()
# p_coverage_5to3_spikein

p_coverage_5to3 <- ggplot(plotData[cellLine %in% cellLines], aes(x = pos_bin, y = normCount, group = pc))+
    geom_line(aes(col = protocol_type_factor), alpha = 0.1)+
    geom_line(data = plotData[cellLine == "A549"], aes(x = pos_bin, y = mean_count,group = pc, col = protocol_type_factor))+
     # geom_ribbon(aes(ymin=pmax(0,cumEstQ1)*100, ymax=pmin(1,cumEstQ3)*100), col = NA, alpha=0.2,show.legend = FALSE)+
    scale_x_continuous(limits = c(0,100),
                       breaks = c(0,100),
                     labels = c("5'","3'"))+
    scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec[1:5],
                       labels = protocolLabel[1:5])+
    ylab("Coverage")+
    xlab("Transcription direction")+
    theme_classic()
p_coverage_5to3
```


## d full-length
```{r, eval = FALSE}
seOutput <- readRDS("bambuOutput_May25.rds")
distTablesList <- metadata(seOutput)$distTables
# check for short read 
type_dist <- lapply(seq_along(distTablesList), function(k){
  print(k)
  reads <- data.table(as.data.frame(distTablesList[[k]]))
  reads[, multi := .N, by = readClassId]
  reads[, type := ifelse(multi==1&all(equal), "FIM",
                  ifelse(multi==1&(!all(equal)), "SIM",
                  ifelse(multi>1&(all(equal)),"MFIM",
                  ifelse(multi>1&(!all(equal))&(any(equal)),"FSIM",
                  ifelse(multi>1&(all(!equal)),"MSIM",NA))))), by = readClassId]
  reads <- unique(reads[,.(readClassId, readCount, type)], by = NULL)
  
  reads_type_sum <- unique(reads[, list(read_count = sum(readCount)), by = type], by = NULL)
  reads_type_sum[, nTotal := sum(read_count)]
  reads_type_sum[, runname := names(distTablesList)[k]]
  return(reads_type_sum)
})
saveRDS(type_dist, file = "type_dist_lr.rds")
```



```{r, eval = FALSE}
seOutputPacBio <- readRDS("bambuOutput_PacBio_May22.rds")
distTablesList <- metadata(seOutputPacBio)$distTables
# check for short read 
type_dist <- lapply(seq_along(distTablesList), function(k){
  print(k)
  reads <- data.table(as.data.frame(distTablesList[[k]]))
  reads[, multi := .N, by = readClassId]
  reads[, type := ifelse(multi==1&all(equal), "FIM",
                  ifelse(multi==1&(!all(equal)), "SIM",
                  ifelse(multi>1&(all(equal)),"MFIM",
                  ifelse(multi>1&(!all(equal))&(any(equal)),"FSIM",
                  ifelse(multi>1&(all(!equal)),"MSIM",NA))))), by = readClassId]
  reads <- unique(reads[,.(readClassId, readCount, type)], by = NULL)
  
  reads_type_sum <- unique(reads[, list(read_count = sum(readCount)), by = type], by = NULL)
  reads_type_sum[, nTotal := sum(read_count)]
  reads_type_sum[, runname := names(distTablesList)[k]]
  return(reads_type_sum)
})
saveRDS(type_dist, file = "type_dist_lr_pacbio.rds")
```




```{r, eval = FALSE}
seSpikein <- readRDS("bambuOutput_spikein_bam_May22.rds")
distTablesList <- metadata(seSpikein)$distTables
# check for short read 
type_dist <- lapply(seq_along(distTablesList), function(k){
  print(k)
  reads <- data.table(as.data.frame(distTablesList[[k]]))
  reads[, multi := .N, by = readClassId]
  reads[, type := ifelse(multi==1&all(equal), "FIM",
                  ifelse(multi==1&(!all(equal)), "SIM",
                  ifelse(multi>1&(all(equal)),"MFIM",
                  ifelse(multi>1&(!all(equal))&(any(equal)),"FSIM",
                  ifelse(multi>1&(all(!equal)),"MSIM",NA))))), by = readClassId]
  reads <- unique(reads[,.(readClassId, readCount, type)], by = NULL)
  
  reads_type_sum <- unique(reads[, list(read_count = sum(readCount)), by = type], by = NULL)
  reads_type_sum[, nTotal := sum(read_count)]
  reads_type_sum[, runname := names(distTablesList)[k]]
  return(reads_type_sum)
})
saveRDS(type_dist, file = "type_dist_spikein.rds")
```


```{r, eval = FALSE}
sr_se_files <- dir(".", pattern = ".rds", full.names = TRUE)
# check for short read 
type_dist_sr <- lapply(sr_se_files, function(b){
  print(b)
  rname <- basename(b)
  sr_se <- readRDS(b)
  reads <- data.table(as.data.frame(metadata(sr_se)$distTables[[1]]))
  reads[, multi := .N, by = readClassId]
  reads[, type := ifelse(multi==1&all(equal), "FIM",
                  ifelse(multi==1&(!all(equal)), "SIM",
                  ifelse(multi>1&(all(equal)),"MFIM",
                  ifelse(multi>1&(!all(equal))&(any(equal)),"FSIM",
                  ifelse(multi>1&(all(!equal)),"MSIM",NA))))), by = readClassId]
  reads <- unique(reads[,.(readClassId, readCount, type)], by = NULL)
  
  reads_type_sum <- unique(reads[, list(read_count = sum(readCount)), by = type], by = NULL)
  reads_type_sum[, nTotal := sum(read_count)]
  reads_type_sum[, runname := basename(b)]
  return(reads_type_sum)
})
saveRDS(type_dist_sr, file = "type_dist_sr.rds")
```




```{r}
type_dist_lr <- readRDS("type_dist_lr.rds")#_protein_coding
type_dist_lr_pacbio <- readRDS("type_dist_lr_pacbio.rds")

# type_dist_lr <- readRDS("type_dist_lr_spliced_reads_only_protein_coding.rds")#_protein_coding
# type_dist_lr_pacbio <- readRDS("type_dist_lr_pacbio__spliced_reads_only_protein_coding.rds")
type_dist_sr <- readRDS("type_dist_sr.rds")
type_dist_spikein <- readRDS("type_dist_spikein.rds")
type_dist_spikein_sr <- readRDS("type_dist_spikein.rds")
type_dist_spikein_sr <- type_dist_spikein_sr[grep("Illumina", unlist(lapply(type_dist_spikein_sr, function(x) unique(x$runname))))]
type_dist <- rbindlist(c(type_dist_lr,type_dist_sr,type_dist_spikein, type_dist_spikein_sr, type_dist_lr_pacbio), fill = TRUE)
type_dist[, runname := gsub("GIS_Hct116_PacBio-SMRTcell_Rep5_Run1","GIS_HCT116_PacBio-SMRTcell_Rep7_Run1",gsub("HCT116","Hct116",gsub("_genome.rds|STAR_alignment|_sorted","", runname)))]
type_dist[, old_runname := runname]
temp <- samplesRC_combined[,.(old_runname, runname)]
setnames(temp, "runname", "new_name")
type_dist <- temp[type_dist, on = "old_runname"]
type_dist[!is.na(new_name), runname := new_name]
type_dist[, `:=`(old_runname = NULL, new_name = NULL)]
type_dist <- samplesRC_combined[type_dist, on = "runname"]
type_dist[, `:=`(protocol_type = gsub("-SMRTcell","",protocol_type))]
type_dist[, protocol_type_factor := factor(protocol_type, c('directRNA','directcDNA','cDNA','PacBio','Illumina'),
                                                    protocolVec)]


# sum_stats <- unique(type_dist[, list(full_length = sum(.SD[type %in% c("FIM","FSIM","MFIM")]$read_count)), by = list(protocol_type_factor,runname, nTotal)], by = NULL)
# median(sum_stats[protocol_type_factor == "directRNA"]$full_length/sum_stats[protocol_type_factor == "directRNA"]$nTotal)
# median(sum_stats[protocol_type_factor == "directcDNA"]$full_length/sum_stats[protocol_type_factor == "directcDNA"]$nTotal)
# median(sum_stats[protocol_type_factor == "cDNA"]$full_length/sum_stats[protocol_type_factor == "cDNA"]$nTotal)
# median(sum_stats[protocol_type_factor == "PacBio"]$full_length/sum_stats[protocol_type_factor == "PacBio"]$nTotal)

p_full_length <- ggplot(type_dist[!grepl("all",runname)], aes(x = type, y = read_count/nTotal*100))+
    #geom_violin(draw_quantiles = NULL, trim = TRUE,adjust = 0.5)+
    geom_boxplot(aes(fill = protocol_type_factor,col = protocol_type_factor), width = 0.5, position=position_dodge(0.75), outlier.shape = NA)+
    scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec[1:5],
                       labels = protocolLabel[1:5])+
    ylab("Percent of reads")+
    scale_x_discrete(limits = c("FIM","SIM","MSIM","FSIM","MFIM"),
                     label = c("Unique\n(Full-length)",
                               "Unique\n(Partial)",
                               "Multi-mapping\n(Partial)",
                               "Multi-mapping\n(Full-length/partial)",
                               "Multi-mapping\n(Full-length)"))+
    xlab("Alignment type")+
    theme_classic()
p_full_length
# pdf("full_length_spliced_reads_protein_coding_genesonly.pdf", width = 6, height = 9/2)
# p_full_length + theme(legend.position = "top")
# dev.off()
```




## e fraction of transcription
```{r 6-transcript-diversity}
txLengths.tbldf <- general_list$txLengths.tbldf
# seSpikein <- readRDS("bambuOutput_spikein_bam.rds")
# seOutput <- readRDS("bambuOutput_spikein_bam.rds")
com_data_gene <- readRDS("combinedExpressionDataGene_19June2023.rds")
com_data_gene[, protocol_general := gsub("-SMRTcell","", protocol_general)]
geneTable <- unique(txLengths.tbldf[,.(gene_id, gene_biotype,hgnc_symbol)])
setnames(geneTable, "gene_id","gene_name")
#com_data_gene_tmp <- geneTable[com_data_gene[grepl("ENSG",gene_name)&(method %in% c("bambu_lr","salmon_sr"))], on = "gene_name"]
com_data_gene_tmp <- geneTable[com_data_gene[grepl("ENSG",gene_name)&(method %in% c("salmon_lr","salmon_sr"))], on = "gene_name"]


## rank by protocol and cell line 
## gene order defined for each cell line protocol combination
geneOrder <- com_data_gene_tmp[,.(gene_name,cellLine,normEst,protocol_general,runname)] #cellLine %in% cellLines
geneOrder[, aveExp:=mean(normEst,na.rm=TRUE), by = list(gene_name, protocol_general,cellLine)]
geneOrder <- unique(geneOrder[,.(aveExp, gene_name, protocol_general,cellLine)])
geneOrder[, geneRank:=rank(-aveExp), by = list(protocol_general,cellLine)]
com_data_gene_tmp <- com_data_gene_tmp[geneOrder, on = c("gene_name","protocol_general","cellLine")]


dt <- do.call("rbind",lapply(unique(com_data_gene_tmp$runname), function(r){
    dt <- com_data_gene_tmp[runname == r]
   
    dt[order(geneRank), cumEst:=cumsum(normEst)/sum(normEst)]
    dt <- dt[,.(geneRank,cumEst,runname, protocol_general, gene_biotype, aveExp, cellLine)] 
    return(dt)
}))

library(ggrepel)

dt[, cell_protocol := paste0(cellLine, protocol_general)]
dt[, protocol_reduced := gsub("RandomPrimer","",protocol_general)]



dt_agg <- dt[, list(geneRank_new = 1:max(ceiling(geneRank)),
                    cumEstFun = approxfun(geneRank, cumEst)(1:max(ceiling(geneRank)))), by = list(protocol_reduced, runname)]


dt_agg_ave <- unique(dt_agg[!grepl("Myeloma",runname)&!grepl("RandomPrimer",runname), list(cumEstAve = median(cumEstFun,na.rm = TRUE), cumEstQ1 = quantile(cumEstFun,na.rm = TRUE, prob = 0.25), cumEstQ3 = quantile(cumEstFun,na.rm = TRUE, prob = 0.75)), by = list(protocol_reduced, geneRank_new)])

```


```{r}
p_transcript_diversity <- ggplot(dt_agg_ave, aes(x = geneRank_new, y = cumEstAve*100, group = protocol_reduced, col = protocol_reduced, fill = protocol_reduced))+
   
    geom_ribbon(aes(ymin=pmax(0,cumEstQ1)*100, ymax=pmin(1,cumEstQ3)*100), col = NA, alpha=0.2,show.legend = FALSE)+
    geom_line()+
    geom_vline(xintercept = 1000, linetype = 2)+
   scale_x_log10()+
     scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec,
                       labels = protocolLabel)+
    scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
   geom_text_repel(data = dt_agg_ave[geneRank_new == 1000],aes(label =  paste0(round(cumEstAve*100),"%")), size = 3, nudge_x = -1)+
    ylab("Percent of total transcription (%)")+
    xlab("Number of genes (rank by expression)")+
    theme_classic()
p_transcript_diversity

# # for salmon_lr
# pdf("p_gene_diversity_salmon.pdf", width = 6, height = 4)
# p_transcript_diversity
# dev.off()
```


## f cDNA vs dRNA bias 
```{r 9-pcr-bias-2}
# resOld <- readRDS("pcr_bias_results.rds")
data <- readRDS("bambuOutput_May25.rds")
txLengths <- sum(width(rowRanges(data)))
geneLengths <- tapply(txLengths, rowData(data)$GENEID, min)
geneLengths.max <- tapply(txLengths, rowData(data)$GENEID, max)

res <- readRDS("pcr_bias_results_short_read_updated.rds")
match_names <- res[[5]]
protRes <- gsub('.*(_|d|D)(irectcDNA|cDNA|irectRNA|Illumina)(Stranded|_|R).*','\\2', match_names)
cellLines <- "all"
cellLineRes <- unlist(lapply(strsplit(gsub("STAR_alignment","",match_names),"_"),"[[",2))
resRelCov = (res[[4]]/res[[1]])

scoreList <- lapply(cellLines, function(k){
if(k == "Spikein"){
    meanCount.cdna <- res[[1]][,protRes=='cDNA'&(cellLineRes %in% k)]
meanCount.dcdna <- res[[1]][,protRes=='irectcDNA'&(cellLineRes %in% k)]
meanCount.drna <-  res[[1]][,protRes=='irectRNA'&(cellLineRes %in% k)]
meanCount.illumina <-  res[[1]][,protRes=='Illumina'&(cellLineRes %in% k)]
mean.cdna <-  resRelCov[,protRes=='cDNA'&(cellLineRes %in% k)]
mean.dcdna <-  resRelCov[,protRes=='irectcDNA'&(cellLineRes %in% k)]
mean.drna <-  resRelCov[,protRes=='irectRNA'&(cellLineRes %in% k)]
mean.illumina <-  resRelCov[,protRes=='Illumina'&(cellLineRes %in% k)]
}else if(k == "all"){
    meanCount.cdna <- rowMeans(res[[1]][,protRes=='cDNA'],na.rm=TRUE)
meanCount.dcdna <- rowMeans(res[[1]][,protRes=='irectcDNA'],na.rm=TRUE)
meanCount.drna <- rowMeans(res[[1]][,protRes=='irectRNA'],na.rm=TRUE)
meanCount.illumina <- rowMeans(res[[1]][,protRes=='Illumina'],na.rm=TRUE)
mean.cdna <- rowMeans(resRelCov[,protRes=='cDNA'],na.rm=TRUE)
mean.dcdna <- rowMeans(resRelCov[,protRes=='irectcDNA'],na.rm=TRUE)
mean.drna <- rowMeans(resRelCov[,protRes=='irectRNA'],na.rm=TRUE)
mean.illumina <- rowMeans(resRelCov[,protRes=='Illumina'],na.rm=TRUE)
}else{
     meanCount.cdna <- rowMeans(res[[1]][,protRes=='cDNA'&(cellLineRes %in% k)],na.rm=TRUE)
meanCount.dcdna <- rowMeans(res[[1]][,protRes=='irectcDNA'&(cellLineRes %in% k)],na.rm=TRUE)
meanCount.drna <- rowMeans(res[[1]][,protRes=='irectRNA'&(cellLineRes %in% k)],na.rm=TRUE)
meanCount.illumina <- rowMeans(res[[1]][,protRes=='Illumina'&(cellLineRes %in% k)],na.rm=TRUE)
mean.cdna <- rowMeans(resRelCov[,protRes=='cDNA'&(cellLineRes %in% k)],na.rm=TRUE)
mean.dcdna <- rowMeans(resRelCov[,protRes=='irectcDNA'&(cellLineRes %in% k)],na.rm=TRUE)
mean.drna <- rowMeans(resRelCov[,protRes=='irectRNA'&(cellLineRes %in% k)],na.rm=TRUE)
mean.illumina <- rowMeans(resRelCov[,protRes=='Illumina'&(cellLineRes %in% k)],na.rm=TRUE)
}

setBiasCandidates =  meanCount.cdna>30 & meanCount.dcdna>30 & meanCount.drna>30 & meanCount.illumina>30
score.cdna.dcdna=(mean.cdna-mean.dcdna)/sqrt((mean.cdna*(1-mean.cdna))/meanCount.cdna+(mean.dcdna*(1-mean.dcdna))/meanCount.dcdna)
score.cdna.dcdna.p.adjust <- p.adjust(pnorm(abs(score.cdna.dcdna), lower.tail = F), method = 'fdr')
score.cdna.drna=(mean.cdna-mean.drna)/sqrt((mean.cdna*(1-mean.cdna))/meanCount.cdna+(mean.drna*(1-mean.drna))/meanCount.drna)
score.cdna.drna.p.adjust <- p.adjust(pnorm(abs(score.cdna.drna), lower.tail = F), method = 'fdr')
score.dcdna.drna=(mean.dcdna-mean.drna)/sqrt((mean.dcdna*(1-mean.dcdna))/meanCount.dcdna+(mean.drna*(1-mean.drna))/meanCount.drna)
score.dcdna.drna.p.adjust <- p.adjust(pnorm(abs(score.dcdna.drna), lower.tail = F), method = 'fdr')
score.dcdna.illumina=(mean.dcdna-mean.illumina)/sqrt((mean.dcdna*(1-mean.dcdna))/meanCount.dcdna+(mean.illumina*(1-mean.illumina))/meanCount.illumina)
score.dcdna.illumina.p.adjust <- p.adjust(pnorm(abs(score.dcdna.illumina), lower.tail = F), method = 'fdr')
score.cdna.illumina=(mean.cdna-mean.illumina)/sqrt((mean.cdna*(1-mean.cdna))/meanCount.cdna+(mean.illumina*(1-mean.illumina))/meanCount.illumina)
score.cdna.illumina.p.adjust <- p.adjust(pnorm(abs(score.cdna.illumina), lower.tail = F), method = 'fdr')
score.drna.illumina=(mean.drna-mean.illumina)/sqrt((mean.drna*(1-mean.drna))/meanCount.drna+(mean.illumina*(1-mean.illumina))/meanCount.illumina)
score.drna.illumina.p.adjust <- p.adjust(pnorm(abs(score.drna.illumina), lower.tail = F), method = 'fdr')
# smoothScatter(mean.cdna[setBiasCandidates],
#               mean.dcdna[setBiasCandidates], xlim=c(0,1), ylim=c(0,1), nrpoints = 10000)
# abline(0,1)
# points(mean.cdna[score.cdna.dcdna< (-3.5) ], mean.dcdna[score.cdna.dcdna<(-3.5) ], col=2)
# 

dat <- data.table(meanCount.cdna = meanCount.cdna,
                  meanCount.dcdna = meanCount.dcdna,
                  meanCount.drna = meanCount.drna,
                  meanCount.illumina= meanCount.illumina,
                  setBiasCandidates = setBiasCandidates,
                  mean.cdna = mean.cdna,
                  mean.dcdna = mean.dcdna,
                  mean.drna = mean.drna,
                  mean.illumina = mean.illumina,
                  score.cdna.dcdna = score.cdna.dcdna,
                  score.cdna.dcdna.p.adjust = score.cdna.dcdna.p.adjust,
                  score.cdna.drna = score.cdna.drna,
                  score.cdna.drna.p.adjust = score.cdna.drna.p.adjust,
                  score.dcdna.drna = score.dcdna.drna,
                  score.dcdna.drna.p.adjust = score.dcdna.drna.p.adjust,
                  score.dcdna.illumina = score.dcdna.illumina,
                  score.dcdna.illumina.p.adjust = score.dcdna.illumina.p.adjust,
                  score.cdna.illumina = score.cdna.illumina,
                  score.cdna.illumina.p.adjust = score.cdna.illumina.p.adjust,
                  score.drna.illumina = score.drna.illumina,
                  score.drna.illumina.p.adjust = score.drna.illumina.p.adjust,
                  geneID = names(geneLengths.max))
				  
score_dat <- dat[,c(10,12,14,16,18,20), with = FALSE]
return(list(data.table(number_of_genes = c(apply(score_dat<(-3.5),2,sum,na.rm = TRUE),
                                                     apply(score_dat> 3.5,2,sum,na.rm = TRUE)),
           comparison = c("cDNA vs dcDNA", "cDNA vs dRNA", "dcDNA vs dRNA", "dcDNA vs Illumina", "cDNA vs Illumina", "dRNA vs Illumina",
                          "dcDNA vs cDNA", "dRNA vs cDNA", "dRNA vs dcDNA", "Illumina vs dcDNA", "Illumina vs cDNA", "Illumina vs dRNA"),
						  cellLine = k),dat))


})
cellLines <- general_list$cellLines
```
```{r}
library(ggpubr)
# pList <- lapply(1, function(x){ #1:7
dat <- scoreList[[1]][[2]]
p_cdna_drna <- ggplot(data = dat[which(setBiasCandidates)], aes(x = mean.drna, y = mean.cdna))+
    xlim(0,1)+
    ylim(0,1)+
    expand_limits(x = c(0,1), y = c(0,1))+
    xlab("Mean coverage per gene \n (direct RNA)")+
    ylab("Mean coverage per gene \n (cDNA)")+
    geom_hex(aes(fill = stat(count)),
             # stat(cut(log(count), 
             #               breaks = log(c(1, 10, 100, 1000,10000,Inf)), 
             #               labels = F, right = T, include.lowest = T))), 
             binwidth = 0.01) +
    scale_fill_gradient(name = 'count', low = "grey", high = "black")+
    #labels = c('0', '1', '2', '3','4+')
    #)+
    stat_cor(aes(label = ..r.label..),method = "spearman", cor.coef.name = "Sp.R")+
    # geom_point(data = dat[setBiasCandidates&abs(score.cdna.drna) > 3.5],
    #            aes(x = mean.drna, y = mean.cdna, col = score.cdna.drna> 3.5), shape = 1, size = 0.5)+
    theme_classic()

p_cdna_drna
```


## arrane main figure
```{r}
library(ggpubr)
pdf("figure_2_draft_31Aug2023.pdf", width = 10, height = 12)
ggarrange(p_sequencing_depth+theme(legend.position="none"), 
          p_mapped_read_length+theme(legend.position="none"),
          p_coverage_5to3+theme(legend.position="none"), 
          p_full_length+theme(legend.position="none"), 
          p_transcript_diversity+theme(legend.position="none"),
          p_cdna_drna+theme(legend.position="none"),
           as_ggplot(get_legend(p_sequencing_depth)),
          as_ggplot(get_legend(p_coverage_5to3)),
          as_ggplot(get_legend(p_full_length)),
           as_ggplot(get_legend(p_cdna_drna)),
          common.legend = TRUE,labels = c(letters[1:6],rep("",4)),
          ncol = 3, nrow = 4, align = "hv")
dev.off()

# pdf("full_length.pdf", width = 10, height = 4)
# ggarrange(p_full_length, p_number_of_transcripts_per_read, p_njunc,
#           common.legend = TRUE,labels = "auto",
#           ncol = 3, nrow = 1, align = "hv")
# dev.off()
# 
# pdf("transcript_diversity.pdf", width = 4, height = 3)
# p_transcript_diversity
# dev.off()
# 
# 
# pdf("transcript_diversity_atGeneRank1000.pdf", width = 10, height = 12)
# p_gene_diversity
# dev.off()

```

# Supplementary Fig. 1 & 3

## a-b expected sequencing throughput and cost
```{r}
## all price at USD dollar
cost.promethion_2880= 600#per flowcell when buy at 2880 flowcells
cost.promethion_96= 785#per flowcell when buy at 96 flowcells
cost.promethion_12 = 855 # per flowcell when buy at 12 flowcells  
cost.gridion_96 = 475  # per flowcell when buy at 96 flowcells 
cost.gridion_12 = 790 #  per flowcell when buy at 12 flowcells   # 475 per flowcellcost.pacbio = 
# (49955-10*599)/60
cost.illumina = 2200 # 
cost.gridion = 855
cost.promethion = 2480

#Typical number of reads	1 million full-length per flow cell on MinION/GridION; 7 million per flow cell on PromethION
readCount.promethion.directRNA=7 # based on the most recent promehion run 
readCount.gridion.directRNA=1

#5-10 million full-length per flow cell on MinION/GirdION; 15-30 million per flow cell on PromethION
readCount.promethion.directcDNA=30
readCount.gridion.directcDNA=10
readCount.promethion.directcDNA.observed=37

#7–12+ million full-length mapped readsper flow cell on MinION/GridION; 50+ million mapped reads per flow cell on PromethION
readCount.promethion.PCRcDNA=70 # pcs109
readCount.gridion.PCRcDNA=20 # pcs111 20
readCount.promethion.PCRcDNA.observed=135 # pcs110


readCount.illumina = 300 #multiplexed 


## this section to be updated 
cost.shortRead=233
cost.directRNA= 390 ##339 # 599/6 = 100
cost.directcDNA=150
cost.directcDNA.single = 420
cost.PCRcDNA= 140 #63 #100
cost.PCRcDNA.single= 400 #274

cost.multiplex.directcDNA= 450 # 
cost.multiplex.PCRcDNA=  140#224   # the price is for 24 plex 

## cost based on minion + kit 
## cost based on gridion +kit 
## cost based on promethion + kit 

#cost per samples:
library(RColorBrewer)
library(data.table)
library(ggplot2)
data_cost <- data.table(cost = c(
    #(cost.directcDNA*(1:12)+cost.promethion+c(0,rep(cost.multiplex.directcDNA,11)))/(1:12),
                            (c(cost.PCRcDNA.single,cost.PCRcDNA*(2:12))+cost.promethion+c(0,rep(cost.multiplex.PCRcDNA,11)))/(1:12),
                            (cost.directRNA*(1:12)+cost.promethion_12*(1:12))/(1:12),
                            #  (c(cost.PCRcDNA.single,cost.PCRcDNA*(2:12))+cost.promethion_96+c(0,rep(cost.multiplex.PCRcDNA,11)))/(1:12),
                            # (cost.directRNA*(1:12)+cost.promethion_96*(1:12))/(1:12),
                            #  (c(cost.PCRcDNA.single,cost.PCRcDNA*(2:12))+cost.promethion_2880+c(0,rep(cost.multiplex.PCRcDNA,11)))/(1:12),
                            # (cost.directRNA*(1:12)+cost.promethion_2880*(1:12))/(1:12),
                           # (cost.directcDNA*(1:12)+cost.gridion+c(0,rep(cost.multiplex.directcDNA,11)))/(1:12),
                            (c(cost.PCRcDNA.single,cost.PCRcDNA*(2:12))+cost.gridion+c(0,rep(cost.multiplex.PCRcDNA,11)))/(1:12),
                            (cost.directRNA*(1:12)+cost.gridion*(1:12))/(1:12),
                           # (c(cost.PCRcDNA.single,cost.PCRcDNA*(2:12))+cost.gridion_96+c(0,rep(cost.multiplex.PCRcDNA,11)))/(1:12),
                            # (cost.directRNA*(1:12)+cost.gridion_96*(1:12))/(1:12),
                            (cost.shortRead*(1:12)+cost.illumina)/(1:12)),
                        samples = rep(c(1:12),5),
                        #platform = c(rep(c("PromethION (12)","PromethION (96)","PromethION (2880)","GridION (12)","GridION (96)"),each = 24),rep("Illumina",12)),
    platform = c(rep(c("PromethION","GridION"),each = 24),rep("Illumina",12)),
                        protocols = c(rep(rep(c("cDNA","directRNA"),each = 12),2),rep("Illumina",12)))

data_cost[, platform_protocol := paste0(platform,"_", protocols)]
data_cost[, protocol_factor := factor(protocols, levels = c("directRNA",
                                                            #"directcDNA",
                                                            "cDNA","Illumina"))]
# pdf("/mnt/projects/SGNExManuscript/figures/cost_per_sample.pdf",width = 8, height = 6)
p_expected_data_cost <- ggplot(data_cost, aes(x = samples, y = cost))+
    geom_line(aes(group = platform_protocol),,col = "grey")+
    geom_point(aes(col = protocol_factor, shape = platform), size = 2)+
    # scale_color_brewer(type = "qual",palette = 3, name ="Protocols")+
  scale_color_manual(values = protocolCol[c(1,3,5)],
                       name = "Protocol",
                       limits = protocolVec[c(1,3,5)],
                       labels = protocolLabel[c(1,3,5)])+
    scale_shape_manual(values = c(16,17,15), 
                       breaks = c("PromethION","GridION","Illumina"),#c("PromethION","GridION","Illumina"), 
                       name = "Platform")+
    xlab("Number of samples")+
    ylab("Cost per sample (USD)")+
    scale_x_continuous(breaks = (1:6)*2)+
    scale_y_continuous(breaks =((1:8)-1)*500, limits = c(0,3000))+
    theme_classic()
# dev.off()

# throughput in giga base 
data_readcount <- data.table(bp = c(
    #readCount.promethion.directcDNA/(1:12)*0.9,
                                      readCount.promethion.PCRcDNA.observed/(1:12)*700,
                                      readCount.promethion.directRNA*rep(1,12)*1000,
                                      #readCount.gridion.directcDNA/(1:12)*0.9,
                                      readCount.gridion.PCRcDNA/(1:12)*700, ## cDNA 
                                      readCount.gridion.directRNA*rep(1,12)*1000,
                                      readCount.illumina/(1:12)*300,
                                      5/1:12*3000),
                        samples = rep(c(1:12),6),
                        platform = c(rep(c("PromethION","GridION"),each = 24),rep("Illumina",12),rep("PacBio",12)),
                        protocols = c(rep(rep(c(
                            #"directcDNA",
                                                "cDNA","directRNA"),each = 12),2),rep("Illumina",12),rep("PacBio",12)))

data_readcount[, platform_protocol := paste0(platform,"_", protocols)]
data_readcount[, protocol_factor := factor(protocols, levels = c("directRNA",
                                                                 #"directcDNA",
                                                                 "cDNA","Illumina","PacBio"))]
# pdf("sequencing_throughput_per_sample.pdf",width = 8, height = 6)
p_expected_throughput <- ggplot(data_readcount, aes(x = samples, y = bp*1000000/1024^3))+
    geom_line(aes(group = platform_protocol),,col = "grey")+
    geom_point(aes(col = protocol_factor, shape = platform), size = 2)+
    # scale_color_brewer(type = "qual",palette = 3, name ="Protocols")+
  scale_color_manual(values = protocolCol[-2],
                       name = "Protocol",
                       limits = protocolVec[-2],
                       labels = protocolLabel[-2])+
    scale_shape_manual(values = c(19,17,15,18), breaks = c("PromethION","GridION","Illumina","PacBio"), name = "Platform")+
    xlab("Number of samples")+
    ylab("Sequencing throughput per sample (million)")+
    scale_x_continuous(breaks = (1:6)*2)+
    # scale_y_continuous(breaks =c(0.3,1,10,100), labels =c(0.01,0.1,1,10,100), limits = c(0.01,120))+
    theme_classic()
# dev.off()
p_expected_data_cost
p_expected_throughput
```
## c spike-in number of reads
```{r}
plotdata <- samplesRC_combined[,.(protocol_type_factor, runname, total_reads, Platform, demultiplexed,cellLine)]
plotdata[is.na(demultiplexed), demultiplexed := FALSE]
## all de-multiplexed are gridion/minion

library(tidyverse)
#library(see)
source("https://raw.githubusercontent.com/datavizpyr/data/master/half_flat_violinplot.R")

# short read are all demultiplexed 
plotdata[Platform == "Illumina", demultiplexed := TRUE]
plotdata[, special_prep:=ifelse(grepl("allSpik",runname),4,ifelse(demultiplexed,3,ifelse(Platform == "PromethION",2,1)))]

# remove spike-in put it in supplementary 
p_sequencing_depth_spikein <- ggplot(plotdata[special_prep == 4], aes(x = protocol_type_factor, y = total_reads/1000000))+
    geom_flat_violin( aes(fill = protocol_type_factor, col = protocol_type_factor),
        alpha = 0.5,
                      position=position_nudge(0.25), 
                      adjust = 2)+
    geom_boxplot( aes(fill = protocol_type_factor), 
                 width = 0.05, outlier.alpha = 0,position=position_nudge(0.25))+
     geom_point(aes(col = protocol_type_factor,shape = as.factor(cellLine)),
               position = position_jitter(0.2),  
               alpha = 0.8)+
     scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec,
                       labels = protocolLabel)+
    scale_shape_manual(values = c(1,0,2,3),
                       name = "Spike-in")+
    xlab("Protocols")+
    ylab(bquote('Number of reads '(10^6)))+ #\nPromethION runs excluded
    scale_y_log10(breaks=c(0.001,0.01,0.1,0.5,1, 3, 5, 10, 20, 
                           40,100),
                  labels =c(0.001,0.01,0.1,0.5,1, 3, 5, 10, 20,
                            40,100))+
    theme_classic()
print(p_sequencing_depth_spikein)
```

## d spike-in read length
```{r}
p_mapped_read_length_spikein <- ggplot(rlData_mat[special_prep == 4], 
    aes(x = protocol_type_factor, y = aveLen))+ #/1000
       geom_flat_violin( aes(fill = protocol_type_factor, col = protocol_type_factor),
        alpha = 0.5,
                      position=position_nudge(0.25), 
                      adjust = 2)+
    geom_boxplot( aes(fill = protocol_type_factor), 
                 width = 0.05, outlier.alpha = 0,position=position_nudge(0.25))+
      geom_point(aes(col = protocol_type_factor,shape = as.factor(cellLine)),
               position = position_jitter(0.2),  
               alpha = 0.8)+
     scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec,
                       labels = protocolLabel)+
     scale_shape_manual(values = c(1,0,2,3),
                       name = "Spike-in")+
    xlab("Protocols")+
    scale_y_log10()+
    ylab("Mean read length")+
    theme_classic()
p_mapped_read_length_spikein
```



## e mapping bases
```{r}
lr_bases <- readRDS("lr_bases_May6.rds")
lr_pacbio_bases <- readRDS("lr_pacbio_bases.rds")
sr_bases <- readRDS("sr_bases.rds")
setnames(sr_bases, "runname", "old_runname")
sr_bases <- samples_wSpikein[,.(runname, old_runname)][sr_bases, on = "old_runname"]
sr_bases[, old_runname := NULL]
spikein_bases_sr <- readRDS("spikein_bases.rds")
spikein_bases_sr[, runname := gsub("\\,","",runname)]
spikein_bases_sr <- spikein_bases_sr[grepl("Illumina",runname)]
spikein_bases_lr <- readRDS("spikein_bases_May5.rds")
spikein_bases_lr[, runname := gsub("sequinMixAv1.0_PacBio","sequinMixAV2_PacBio",runname)]
bases_data <- rbindlist(list(lr_bases, lr_pacbio_bases,sr_bases, spikein_bases_sr, spikein_bases_lr))

bases_data <- rbind(samples[,.(runname, protocol_type, Platform, demultiplexed, cellLine)],
                    samples_wSpikein[grepl("Illumina",runname),.(runname, protocol_type, Platform, demultiplexed, cellLine)])[bases_data , on = "runname"]
bases_data <- bases_data[!is.na(protocol_type)]
bases_data[, protocol_type := gsub("-SMRTcell","",protocol_type)]
#bases_data <- bases_data[cellLine %in% cellLines]
bases_data[, protocol_type_factor := factor(protocol_type, c('directRNA','directcDNA','cDNA',"PacBio","Illumina"),
                                              protocolVec)]

plotdata <- bases_data[,.(protocol_type_factor, runname, cigarmapped_bases, Platform, demultiplexed,cellLine)]
plotdata[is.na(demultiplexed), demultiplexed := FALSE]
plotdata[Platform == "Illumina", demultiplexed := TRUE]
plotdata[, special_prep:=ifelse(grepl("allSpik",runname),4,ifelse(demultiplexed,3,ifelse(Platform == "PromethION",2,1)))]
plotdata[is.na(special_prep), special_prep := 1]
p_total_bases <- ggplot(plotdata[special_prep != 4], aes(x = protocol_type_factor, y = cigarmapped_bases/1024^3))+
    #, color = variable
    geom_flat_violin( aes(fill = protocol_type_factor, col = protocol_type_factor),
        alpha = 0.5,
                      position=position_nudge(0.25), 
                      adjust = 2)+
    geom_boxplot( aes(fill = protocol_type_factor), 
                 width = 0.05, outlier.alpha = 0,position=position_nudge(0.25))+
     geom_point(aes(col = protocol_type_factor,shape = as.factor(special_prep)),
               position = position_jitter(0.2),  
               alpha = 0.8)+
     scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec,
                       labels = protocolLabel)+
    scale_shape_manual(values = c(1,0,2,3),labels = c("MinION/GridION","PromethION","Demultiplexed","Spikein"),
                       name = "Protocol")+
    xlab("Protocols")+
    ylab("Cigar mapped bases(Gb)")+ 
    scale_y_log10(breaks=c(0.01,0.1,0,0.5,1,5,10,50),
                  labels =c(0.01,0.1,0,0.5,1,5,10,50))+
    theme_classic()
print(p_total_bases)
```

## f spike-in mapping bases
```{r}
p_total_bases_spikein <- ggplot(plotdata[special_prep == 4], aes(x = protocol_type_factor, y = cigarmapped_bases/1024^3))+
    #, color = variable
    geom_flat_violin( aes(fill = protocol_type_factor, col = protocol_type_factor),
        alpha = 0.5,
                      position=position_nudge(0.25), 
                      adjust = 2)+
    geom_boxplot( aes(fill = protocol_type_factor), 
                 width = 0.05, outlier.alpha = 0,position=position_nudge(0.25))+
     geom_point(aes(col = protocol_type_factor,shape = as.factor(cellLine)),
               position = position_jitter(0.2),  
               alpha = 0.8)+
     scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec,
                       labels = protocolLabel)+
    scale_shape_manual(values = c(1,0,2,3),name = "Spike-in")+
    xlab("Protocols")+
    ylab("Cigar mapped bases(Gb)")+ 
    scale_y_log10(breaks=c(0.01,0.1,0,0.5,1,5,10,50),
                  labels =c(0.01,0.1,0,0.5,1,5,10,50))+
    theme_classic()
print(p_total_bases_spikein)
```



## g error rate
```{r}
plotdata <- bases_data[,.(protocol_type_factor, runname, error_rate, Platform, demultiplexed,cellLine)]
plotdata[is.na(demultiplexed), demultiplexed := FALSE]
plotdata[Platform == "Illumina", demultiplexed := TRUE]
plotdata[, special_prep:=ifelse(grepl("allSpik",runname),4,ifelse(demultiplexed,3,ifelse(Platform == "PromethION",2,1)))]
plotdata[is.na(special_prep), special_prep := 1]
p_error_rate <- ggplot(plotdata[special_prep != 4], aes(x = protocol_type_factor, y = error_rate))+
    #, color = variable
    geom_flat_violin( aes(fill = protocol_type_factor, col = protocol_type_factor),
        #fill = "lightblue",col = "lightblue", 
        alpha = 0.5,
                      position=position_nudge(0.25), 
                      adjust = 2)+#color = "steelblue",
    #geom_flat_violin(draw_quantiles = NULL, trim = TRUE,  adjust = 0.5)+#color = "steelblue",
    geom_boxplot( aes(fill = protocol_type_factor), #fill = "lightblue",
                 width = 0.05, outlier.alpha = 0,position=position_nudge(0.25))+#color = "steelblue",
    #             geom="pointrange", position = position_jitterdodge(0.5))+
    #geom_jitter(width = 0.2,alpha = 0.5)+#color = "steelblue",
   geom_point(aes(col = protocol_type_factor,shape = as.factor(special_prep)),
                #binaxis = "y", stackdir = "centerwhole",
                # right = FALSE,
                #fill = "darkgrey",binwidth = 0.1,# col = cellLine),
               position = position_jitter(0.2),  
               # dotsize = 1,
              #col = "darkgrey", 
              # size = 2,
               alpha = 0.8)+
     scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec,
                       labels = protocolLabel)+
    # scale_color_brewer(type = "qual", palette = 2,
    #                    limits = c("sequinMixAV1","sequinMixAV2","SIRV-1","SIRV-4"),
    #                    labels = c("Sequin MixA V1","Sequin MixA V2","SIRV-1","SIRV-4"))+
     scale_shape_manual(values = c(1,0,2,3),labels = c("MinION/GridION","PromethION","Demultiplexed","Spikein"),
                       name = "Protocol")+
    # scale_y_log10()+
    xlab("Protocols")+
    ylab("Error rate")+ #\nPromethION runs excluded
    theme_classic()
print(p_error_rate)
```
```{r}
p_error_rate_log10 <- ggplot(plotdata[special_prep != 4], aes(x = protocol_type_factor, y = error_rate))+
    #, color = variable
     geom_flat_violin( aes(fill = protocol_type_factor, col = protocol_type_factor),
        #fill = "lightblue",col = "lightblue", 
        alpha = 0.5,
                      position=position_nudge(0.25), 
                      adjust = 2)+#color = "steelblue",
    #geom_flat_violin(draw_quantiles = NULL, trim = TRUE,  adjust = 0.5)+#color = "steelblue",
    geom_boxplot( aes(fill = protocol_type_factor), #fill = "lightblue",
                 width = 0.05, outlier.alpha = 0,position=position_nudge(0.25))+#color = "steelblue",
    #             geom="pointrange", position = position_jitterdodge(0.5))+
    #geom_jitter(width = 0.2,alpha = 0.5)+#color = "steelblue",
   geom_point(aes(col = protocol_type_factor,shape = as.factor(special_prep)),
                #binaxis = "y", stackdir = "centerwhole",
                # right = FALSE,
                #fill = "darkgrey",binwidth = 0.1,# col = cellLine),
               position = position_jitter(0.2),  
               # dotsize = 1,
              #col = "darkgrey", 
              # size = 2,
               alpha = 0.8)+
     scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec,
                       labels = protocolLabel)+
    # scale_color_brewer(type = "qual", palette = 2,
    #                    limits = c("sequinMixAV1","sequinMixAV2","SIRV-1","SIRV-4"),
    #                    labels = c("Sequin MixA V1","Sequin MixA V2","SIRV-1","SIRV-4"))+
    scale_shape_manual(values = c(1,0,2,3),labels = c("MinION/GridION","PromethION","Demultiplexed","Spikein"),
                       name = "Protocol")+
    scale_y_log10()+
    xlab("Protocols")+
    ylab("Error rate")+ #\nPromethION runs excluded
    theme_classic()
print(p_error_rate_log10)
```

## h spikein-error rate
```{r}
p_error_rate_log10_spikein <- ggplot(plotdata[special_prep == 4], aes(x = protocol_type_factor, y = error_rate))+
    #, color = variable
     geom_flat_violin( aes(fill = protocol_type_factor, col = protocol_type_factor),
        #fill = "lightblue",col = "lightblue", 
        alpha = 0.5,
                      position=position_nudge(0.25), 
                      adjust = 2)+#color = "steelblue",
    #geom_flat_violin(draw_quantiles = NULL, trim = TRUE,  adjust = 0.5)+#color = "steelblue",
    geom_boxplot( aes(fill = protocol_type_factor), #fill = "lightblue",
                 width = 0.05, outlier.alpha = 0,position=position_nudge(0.25))+#color = "steelblue",
    #             geom="pointrange", position = position_jitterdodge(0.5))+
    #geom_jitter(width = 0.2,alpha = 0.5)+#color = "steelblue",
   geom_point(aes(col = protocol_type_factor,shape = as.factor(cellLine)),
                #binaxis = "y", stackdir = "centerwhole",
                # right = FALSE,
                #fill = "darkgrey",binwidth = 0.1,# col = cellLine),
               position = position_jitter(0.2),  
               # dotsize = 1,
              #col = "darkgrey", 
              # size = 2,
               alpha = 0.8)+
     scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec,
                       labels = protocolLabel)+
    # scale_color_brewer(type = "qual", palette = 2,
    #                    limits = c("sequinMixAV1","sequinMixAV2","SIRV-1","SIRV-4"),
    #                    labels = c("Sequin MixA V1","Sequin MixA V2","SIRV-1","SIRV-4"))+
    scale_shape_manual(values = c(1,0,2,3),name = "Spike-in")+
    scale_y_log10()+
    xlab("Protocols")+
    ylab("Error rate")+ #\nPromethION runs excluded
    theme_classic()
print(p_error_rate_log10_spikein)
```

## i spike-in coverage 
```{r}
covData <- readRDS(paste0(wkdir,"output_guppy6.4.2/covData.rds"))
covData[grepl("sequinMixAv1.0_PacBio",runname), runname := gsub("v1.0","V2",runname)]
covData <- samplesRC_combined[, .(protocol_type_factor, cellLine,runname)][covData, on = "runname"]
plotData <- unique(covData[, list(aveNorm=mean(ave_bin_count)), by = list(pos_bin,protocol_type_factor, cellLine)])
plotData[, normCount:=aveNorm/max(aveNorm), by = list(protocol_type_factor,cellLine)]


plotData[, mean_count :=mean(normCount), by = list(protocol_type_factor,pos_bin)]
plotData[, min_count := min(normCount), by = list(protocol_type_factor,pos_bin)]
plotData[, max_count := max(normCount), by = list(protocol_type_factor,pos_bin)]

plotData[, pc := paste0(protocol_type_factor," ",cellLine)]
p_coverage_5to3_spikein <- ggplot(plotData[grep("sequin|SIRV",cellLine)], aes(x = pos_bin, y = normCount, group = pc))+
    geom_line(aes(col = protocol_type_factor))+
    scale_x_continuous(limits = c(0,100),
                       breaks = c(0,100),
                     labels = c("5'","3'"))+
    scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec[1:5],
                       labels = protocolLabel[1:5])+
    ylab("Coverage")+
    xlab("Transcription direction")+
    theme_classic()
p_coverage_5to3_spikein
```

## j read coverage  spike-in
```{r 8-read-coverage, eval = FALSE}
rdfile <- c(dir("readLength17Apr/", full.names = TRUE, recursive = TRUE),
            dir("readLength17Apr_pacbio/", full.names = TRUE, recursive = TRUE),
            dir("readLength17Apr_spikein/", full.names = TRUE, recursive = TRUE))

rdfile_name <- gsub("readLength_","",tools::file_path_sans_ext(BiocGenerics::basename(rdfile)))
txLengths.tbldf <- general_list$txLengths.tbldf
txInfo <- txLengths.tbldf[,.(tx_name, nexon, tx_len, nisoform, gene_biotype)]
rlData <- do.call("rbind",lapply(seq_along(rdfile_name), function(r){
    print(r)
        rlData <- readRDS(rdfile[r])
        rlData <- txInfo[rlData, on = "tx_name"]
        rlData[, txCvr:=aligned_len/tx_len]
        rlData <- rlData[!is.na(txCvr)]
        rlData[,runname := rdfile_name[r]]
        rlData[, txCvr_round:=ceiling(txCvr*1000)/1000]
        cvr <- unique(rlData[,list(n=.N), by = list(txCvr_round,
                                                 runname)])
        cvr[order(txCvr_round, decreasing = TRUE), cn:=cumsum(n)/sum(n)]
        return(cvr)
    })
    )
saveRDS(rlData, file = paste0(wkdir,"readCoverage_May6.rds"))
```

```{r 8-read-coverage-2}
rlData <- readRDS(paste0(wkdir,"readCoverage_May6.rds"))
rlData_sr <- readRDS(paste0(wkdir,"readCoverage.rds"))
rlData_sr <- rlData_sr[grepl("Illumina",runname)]
rlData_sr[, runname := gsub("\\,","",runname)]
name_rr <- rbind(samples[grep("Illumina",runname),.(old_runname, runname)],
                 samples_wSpikein[grep("Illumina",runname),.(old_runname, runname)])
setnames(rlData_sr, "runname","old_runname")
rlData_sr <- name_rr[rlData_sr, on = "old_runname"]
rlData_sr[, old_runname:=NULL]
# why illumina spikein so strange 
cvr_dt <- rbind(samples[,.(runname, cellLine, protocol_type)],
                samples_wSpikein[grep("Illumina",runname),.(runname, cellLine, protocol_type)])[rbind(rlData,rlData_sr), on = "runname"]

cvr_agg <- unique(cvr_dt[grepl("allSpik",runname),list(cumEstAve=median(cn),
                                                      cumEstSD = sd(cn)), by = list(protocol_type,cellLine, txCvr_round)])  #cellLine %in% cellLines|

cvr_agg[order(protocol_type, cellLine, txCvr_round), cumsumEstAve:=cumsum(cumEstAve), by = list(protocol_type, cellLine)]
cvr_agg[, cumEstDisp:=cumEstSD/cumsumEstAve]
cvr_agg[, cell_protocol := paste0(cellLine, protocol_type)]
p_read_coverage_spikein <- ggplot(cvr_agg, aes(x = (-txCvr_round), y = cumEstAve*100,group = cell_protocol))+
    # geom_ribbon(aes(ymin=pmax(0,cumEstAve-cumEstDisp), ymax=pmin(1,cumEstAve+cumEstDisp), fill = protocol_type), alpha=0.2,show.legend = FALSE)+
    geom_line(aes(color = protocol_type))+
    scale_x_continuous(breaks = -(5:0)/5,
                       limits = c(-1,0),
                       label = (5:0)*20,expand = c(0,0))+
    scale_y_continuous(expand = c(0,0))+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec,
                       labels = protocolLabel)+
    scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    ylab("Percent of reads")+
    
    xlab("Coverage proportion of transcript(%)")+
    theme_classic()
p_read_coverage_spikein
```


## k read coverage 
```{r}
cvr_agg <- unique(cvr_dt[cellLine %in% cellLines,list(cumEstAve=median(cn),
                                                      cumEstSD = sd(cn)), by = list(protocol_type,cellLine, txCvr_round)])  #cellLine %in% cellLines|

cvr_agg[order(protocol_type, cellLine, txCvr_round), cumsumEstAve:=cumsum(cumEstAve), by = list(protocol_type, cellLine)]
cvr_agg[, cumEstDisp:=cumEstSD/cumsumEstAve]
cvr_agg[, protocol_type := gsub("-SMRTcell","", protocol_type)]
cvr_agg[, cell_protocol := paste0(cellLine, protocol_type)]
p_read_coverage <- ggplot(cvr_agg, aes(x = (-txCvr_round), y = cumEstAve*100,group = cell_protocol))+
    # geom_ribbon(aes(ymin=pmax(0,cumEstAve-cumEstDisp), ymax=pmin(1,cumEstAve+cumEstDisp), fill = protocol_type), alpha=0.2,show.legend = FALSE)+
    geom_line(aes(color = protocol_type))+
    scale_x_continuous(breaks = -(5:0)/5,
                       limits = c(-1,0),
                       label = (5:0)*20,expand = c(0,0))+
    scale_y_continuous(expand = c(0,0))+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec,
                       labels = protocolLabel)+
    scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    ylab("Percent of reads")+
    
    xlab("Coverage proportion of transcript(%)")+
    theme_classic()
p_read_coverage
```

## m transcript diveristy by expression

```{r}
library(ggpubr)
# for salmon
my_comparisons <- list( 
                        c("cDNA", "directRNA"), 
                        c("cDNA", "Illumina"), 
                        c("Illumina", "directRNA"),
                        c("Illumina", "directcDNA"),
                          c("Illumina", "PacBio"),
                        # c("directRNA","directcDNA"),
                        # c("directRNA","PacBio"),
                        c("directcDNA","PacBio"))#)#

# for bambu
# my_comparisons <- list( c("cDNA", "Illumina"), 
#                         c("cDNA", "directRNA"), 
#                         
#                         c("Illumina", "directRNA"),
#                         # c("Illumina", "directcDNA"),
#                         #   c("Illumina", "PacBio"),
#                         c("directRNA","directcDNA"),
#                         c("directRNA","PacBio"),
#                         c("directcDNA","PacBio"))#)#
p_transcript_diversity_boxplot <- ggplot(dt_agg[geneRank_new == 1000], aes(x = reorder(protocol_reduced,cumEstFun),  y = cumEstFun*100))+
    geom_boxplot()+
    stat_compare_means(comparisons = my_comparisons,
                       #label.y = c(1,0.95,0.9,0.85,0.8),
                      label.y = c(1,0.95,0.9,0.85,0.8,0.75)*100)+
                       # method = c("t.test"))+
                       # label.y = c(1,0.95,0.9,0.85,0.8,0.75))+ # Add pairwise comparisons p-value
  # stat_compare_means(label.y = 50) +
    labs(x = "Protocol", y = "Percent of transcription \n for top 1000 ranked genes")+
    theme_classic()
p_transcript_diversity_boxplot
# pdf("/mnt/projects/SGNExManuscript/revisionFigures_guppy6.4.2/p_gene_diversity_boxplot_salmon.pdf", width = 6, height = 4)
# p_transcript_diversity_boxplot
# dev.off()
```
## n transcript diversity by gene length
```{r}
#txLengths.tbldf <- general_list$txLengths.tbldf
# seSpikein <- readRDS("/mnt/projects/SGNExManuscript/output/bambuOutput_spikein_bam.rds")
# seOutput <- readRDS("/mnt/projects/SGNExManuscript/output/bambuOutput_spikein_bam.rds")
# com_data_gene <- readRDS("/mnt/projects/SGNExManuscript/output_guppy6.4.2/combinedExpressionDataGene_25May2023.rds")
# com_data_gene[, protocol_general := gsub("-SMRTcell","", protocol_general)]
geneTable <- unique(txLengths.tbldf[,list(geneLengthMax = max(tx_len)), by = list(gene_id, gene_biotype,hgnc_symbol)])
setnames(geneTable, "gene_id","gene_name")
geneTable[, geneRank := rank(geneLengthMax)]
#com_data_gene_tmp <- geneTable[com_data_gene[grepl("ENSG",gene_name)&(method %in% c("bambu_lr","salmon_sr"))], on = "gene_name"]
com_data_gene_tmp <- geneTable[com_data_gene[grepl("ENSG",gene_name)&(method %in% c("salmon_lr","salmon_sr"))], on = "gene_name"]

## rank by protocol and cell line 
## gene order defined for each cell line protocol combination
# geneOrder <- com_data_gene_tmp[,.(gene_name,cellLine,normEst,protocol_general,runname)] #cellLine %in% cellLines
# geneOrder[, aveExp:=mean(normEst,na.rm=TRUE), by = list(gene_name, protocol_general,cellLine)]
# geneOrder <- unique(geneOrder[,.(aveExp, gene_name, protocol_general,cellLine)])
# geneOrder[, geneRank:=rank(-aveExp), by = list(protocol_general,cellLine)]

# com_data_gene_tmp <- com_data_gene_tmp[geneOrder, on = c("gene_name","protocol_general","cellLine")]


dt <- do.call("rbind",lapply(unique(com_data_gene_tmp$runname), function(r){
    dt <- com_data_gene_tmp[runname == r]
    dt[order(geneRank), cumEst:=cumsum(normEst)/sum(normEst)]
    dt <- dt[,.(geneRank,cumEst,runname, protocol_general, gene_biotype, cellLine)] 
    #dt <- dt[,.(geneRank,cumEst,runname, protocol_general, cellLine)] 
    return(dt)
}))

library(ggrepel)

dt[, cell_protocol := paste0(cellLine, protocol_general)]
dt[, protocol_reduced := gsub("RandomPrimer","",protocol_general)]



dt_agg_geneLength <- dt[, list(geneRank_new = 1:max(ceiling(geneRank)),
                    cumEstFun = approxfun(geneRank, cumEst)(1:max(ceiling(geneRank)))), by = list(protocol_reduced, runname)]
# dt_agg_ave <- unique(dt_agg[!grepl("Myeloma",runname)&!grepl("RandomPrimer",runname), list(cumEstAve = mean(cumEstFun,na.rm = TRUE), cumEstSd = sd(cumEstFun,na.rm = TRUE)), by = list(protocol_reduced, geneRank_new)])
dt_agg_ave_geneLength <- unique(dt_agg_geneLength[!grepl("Myeloma",runname)&!grepl("RandomPrimer",runname), list(cumEstAve = median(cumEstFun,na.rm = TRUE), cumEstQ1 = quantile(cumEstFun,na.rm = TRUE, prob = 0.25), cumEstQ3 = quantile(cumEstFun,na.rm = TRUE, prob = 0.75)), by = list(protocol_reduced, geneRank_new)])
```

# Supplementary Text Fig. 13
```{r}
p_transcript_diversity_by_geneLength <- ggplot(dt_agg_ave_geneLength, aes(x = geneRank_new, y = cumEstAve*100, group = protocol_reduced, col = protocol_reduced, fill = protocol_reduced))+
    #geom_ribbon(aes(ymin=pmax(0,cumEstAve-cumEstSd)*100, ymax=pmin(1,cumEstAve+cumEstSd)*100), col = NA, alpha=0.2,show.legend = FALSE)+
    geom_ribbon(aes(ymin=pmax(0,cumEstQ1)*100, ymax=pmin(1,cumEstQ3)*100), col = NA, alpha=0.2,show.legend = FALSE)+
    geom_line()+
    #geom_line(data = dt_agg[cellLine == "Spikein"], col = "black", size = 3)+
   # geom_vline(xintercept = 1000, linetype = 2)+
    geom_vline(xintercept = 31063, linetype = 2)+
  # scale_x_log10()+
    
     scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec,
                       labels = protocolLabel)+
    scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
   #geom_text_repel(data = dt_agg_ave[geneRank_new == 1000],aes(label =  paste0(round(cumEstAve*100),"%")), size = 3, nudge_x = -1)+
 geom_text_repel(data = dt_agg_ave_geneLength[geneRank_new == 31063],aes(label =  paste0(round(cumEstAve*100),"%")), size = 3, nudge_x = -1000)+
    ylab("Percent of total transcription")+
    xlab("Number of genes (rank by gene length)")+
    theme_classic()
p_transcript_diversity_by_geneLength
# pdf("p_gene_diversity_by_geneLength_salmon.pdf", width = 6, height = 4)
# p_transcript_diversity_by_geneLength
# dev.off()
```

## o transcript diversity by gene length boxplot version with p-value

```{r}
# 31603 1kb
# 46545.5 3kb
library(ggpubr)
my_comparisons <- list( c("cDNA", "directRNA"), c("cDNA", "Illumina"),c("cDNA", "directcDNA"),
                        c("directRNA","directcDNA"),c("Illumina", "directRNA"),c("Illumina", "directcDNA"),c("directcDNA","PacBio"))#,c("directRNA","PacBio"),c("directcDNA","PacBio"))#
p_transcript_diversity_boxplot_by_geneLength <- ggplot(dt_agg_geneLength[geneRank_new == 31063], aes(x = reorder(protocol_reduced,cumEstFun),  y = cumEstFun*100))+
    geom_boxplot()+
    stat_compare_means(comparisons = my_comparisons,
                       #label.y = c(1,0.95,0.9,0.85,0.8),
                      #label.y = c(1,0.95,0.9,0.85,0.8,0.75),
                      label.y = seq(0.4,0.2,length.out = 7)*100)+ #method = c("t.test") by default, stat_compare_means use wilcox.test+
                       # label.y = c(1,0.95,0.9,0.85,0.8,0.75))+ # Add pairwise comparisons p-value
  # stat_compare_means(label.y = 50) +
     labs(x = "Protocol", y = "Percent of transcription \n for genes <= 1kb")+
    theme_classic()

p_transcript_diversity_boxplot_by_geneLength
# pdf("p_gene_diversity_boxplot_by_geneLength_salmon.pdf", width = 6, height = 4)
# p_transcript_diversity_boxplot_by_geneLength
# dev.off()
# p2 <- ggplot(dt_agg[geneRank_new == 46545], aes(x = reorder(protocol_reduced,cumEstFun),  y = cumEstFun))+
#     geom_boxplot()+
#     stat_compare_means(comparisons = my_comparisons,
#                        #label.y = c(1,0.95,0.9,0.85,0.8),
#                       #label.y = c(1,0.95,0.9,0.85,0.8,0.75),
#                       label.y = seq(0.4,0.2,length.out = 7))+ #method = c("t.test") by default, stat_compare_means use wilcox.test+
#                        # label.y = c(1,0.95,0.9,0.85,0.8,0.75))+ # Add pairwise comparisons p-value
#   # stat_compare_means(label.y = 50) +
#     theme_minimal()
```




## l full-length filter by protein coding genes
filter out single exon read classes and those not compatible to protein coding genes 
```{r, eval = FALSE}
ensemblAnnotations.transcripts <- general_list$ensemblAnnotations.transcripts

seOutput <- readRDS("bambuOutput_May25.rds")
distTablesList <- metadata(seOutput)$distTables
# check for short read 
type_dist <- lapply(seq_along(distTablesList), function(k){
  print(k)
  reads <- data.table(as.data.frame(distTablesList[[k]]))
  # filter to use only spliced read class
  reads <- reads[which(!grepl("unspliced", readClassId))]
  reads <- reads[annotationTxId %in% unique(ensemblAnnotations.transcripts[gene_biotype == "protein_coding"]$tx_name)]
  reads[, multi := .N, by = readClassId]
  reads[, type := ifelse(multi==1&all(equal), "FIM",
                  ifelse(multi==1&(!all(equal)), "SIM",
                  ifelse(multi>1&(all(equal)),"MFIM",
                  ifelse(multi>1&(!all(equal))&(any(equal)),"FSIM",
                  ifelse(multi>1&(all(!equal)),"MSIM",NA))))), by = readClassId]
  reads <- unique(reads[,.(readClassId, readCount, type)], by = NULL)
  
  reads_type_sum <- unique(reads[, list(read_count = sum(readCount)), by = type], by = NULL)
  reads_type_sum[, nTotal := sum(read_count)]
  reads_type_sum[, runname := names(distTablesList)[k]]
  return(reads_type_sum)
})
saveRDS(type_dist, file = "type_dist_lr_spliced_reads_only_protein_coding.rds")

## 107
reads_type_sum[, perc := read_count/nTotal]
```


```{r, eval = FALSE}
seOutputPacBio <- readRDS("bambuOutput_PacBio_May22.rds")
distTablesList <- metadata(seOutputPacBio)$distTables
# check for short read 
type_dist <- lapply(seq_along(distTablesList), function(k){
  print(k)
  reads <- data.table(as.data.frame(distTablesList[[k]]))
  # filter to use only spliced read class
  reads <- reads[which(!grepl("unspliced", readClassId))]
  reads <- reads[annotationTxId %in% unique(ensemblAnnotations.transcripts[gene_biotype == "protein_coding"]$tx_name)]
  reads[, multi := .N, by = readClassId]
  reads[, type := ifelse(multi==1&all(equal), "FIM",
                  ifelse(multi==1&(!all(equal)), "SIM",
                  ifelse(multi>1&(all(equal)),"MFIM",
                  ifelse(multi>1&(!all(equal))&(any(equal)),"FSIM",
                  ifelse(multi>1&(all(!equal)),"MSIM",NA))))), by = readClassId]
  reads <- unique(reads[,.(readClassId, readCount, type)], by = NULL)
  
  reads_type_sum <- unique(reads[, list(read_count = sum(readCount)), by = type], by = NULL)
  reads_type_sum[, nTotal := sum(read_count)]
  reads_type_sum[, runname := names(distTablesList)[k]]
  return(reads_type_sum)
})
saveRDS(type_dist, file = "type_dist_lr_pacbio__spliced_reads_only_protein_coding.rds")
```

```{r, eval = FALSE}
sr_se_files <- dir("srSe_revised/", pattern = ".rds", full.names = TRUE)
# check for short read 
type_dist_sr <- lapply(sr_se_files, function(b){
  print(b)
  rname <- basename(b)
  sr_se <- readRDS(b)
  reads <- data.table(as.data.frame(metadata(sr_se)$distTables[[1]]))
  reads[, multi := .N, by = readClassId]
  reads[, type := ifelse(multi==1&all(equal), "FIM",
                  ifelse(multi==1&(!all(equal)), "SIM",
                  ifelse(multi>1&(all(equal)),"MFIM",
                  ifelse(multi>1&(!all(equal))&(any(equal)),"FSIM",
                  ifelse(multi>1&(all(!equal)),"MSIM",NA))))), by = readClassId]
  reads <- unique(reads[,.(readClassId, readCount, type)], by = NULL)
  
  reads_type_sum <- unique(reads[, list(read_count = sum(readCount)), by = type], by = NULL)
  reads_type_sum[, nTotal := sum(read_count)]
  reads_type_sum[, runname := basename(b)]
  return(reads_type_sum)
})
saveRDS(type_dist_sr, file = "type_dist_sr_filtered.rds")
```


```{r}
type_dist_lr <- readRDS("type_dist_lr_spliced_reads_only_protein_coding.rds")#_protein_coding
type_dist_lr_pacbio <- readRDS("type_dist_lr_pacbio__spliced_reads_only_protein_coding.rds")
type_dist_sr <- readRDS("type_dist_sr.rds")
type_dist_spikein <- readRDS("type_dist_spikein.rds")
type_dist_spikein_sr <- readRDS("type_dist_spikein.rds")
type_dist_spikein_sr <- type_dist_spikein_sr[grep("Illumina", unlist(lapply(type_dist_spikein_sr, function(x) unique(x$runname))))]
type_dist <- rbindlist(c(type_dist_lr,type_dist_sr,type_dist_spikein, type_dist_spikein_sr, type_dist_lr_pacbio), fill = TRUE)
type_dist[, runname := gsub("GIS_Hct116_PacBio-SMRTcell_Rep5_Run1","GIS_HCT116_PacBio-SMRTcell_Rep7_Run1",gsub("HCT116","Hct116",gsub("_genome.rds|STAR_alignment|_sorted","", runname)))]
type_dist[, old_runname := runname]
temp <- samplesRC_combined[,.(old_runname, runname)]
setnames(temp, "runname", "new_name")
type_dist <- temp[type_dist, on = "old_runname"]
type_dist[!is.na(new_name), runname := new_name]
type_dist[, `:=`(old_runname = NULL, new_name = NULL)]
type_dist <- samplesRC_combined[type_dist, on = "runname"]
type_dist[, `:=`(protocol_type = gsub("-SMRTcell","",protocol_type))]
type_dist[, protocol_type_factor := factor(protocol_type, c('directRNA','directcDNA','cDNA','PacBio','Illumina'),
                                                    protocolVec)]


# sum_stats <- unique(type_dist[, list(full_length = sum(.SD[type %in% c("FIM","FSIM","MFIM")]$read_count)), by = list(protocol_type_factor,runname, nTotal)], by = NULL)
# median(sum_stats[protocol_type_factor == "directRNA"]$full_length/sum_stats[protocol_type_factor == "directRNA"]$nTotal)
# median(sum_stats[protocol_type_factor == "directcDNA"]$full_length/sum_stats[protocol_type_factor == "directcDNA"]$nTotal)
# median(sum_stats[protocol_type_factor == "cDNA"]$full_length/sum_stats[protocol_type_factor == "cDNA"]$nTotal)
# median(sum_stats[protocol_type_factor == "PacBio"]$full_length/sum_stats[protocol_type_factor == "PacBio"]$nTotal)

p_full_length_filtered <- ggplot(type_dist[!grepl("all",runname)], aes(x = type, y = read_count/nTotal*100))+
    #geom_violin(draw_quantiles = NULL, trim = TRUE,adjust = 0.5)+
    geom_boxplot(aes(fill = protocol_type_factor,col = protocol_type_factor), width = 0.5, position=position_dodge(0.75), outlier.shape = NA)+
    scale_fill_manual(values = protocolCol,
                      name = "Protocol",
                      limits = protocolVec,
                      labels = protocolLabel)+
    scale_color_manual(values = protocolCol,
                       name = "Protocol",
                       limits = protocolVec[1:5],
                       labels = protocolLabel[1:5])+
    ylab("Percent of reads")+
    scale_x_discrete(limits = c("FIM","SIM","MSIM","FSIM","MFIM"),
                     label = c("Unique\n(Full-length)",
                               "Unique\n(Partial)",
                               "Multi-mapping\n(Partial)",
                               "Multi-mapping\n(Full-length/partial)",
                               "Multi-mapping\n(Full-length)"))+
    xlab("Alignment type")+
    theme_classic()
p_full_length_filtered
# pdf("full_length_spliced_reads_protein_coding_genesonly.pdf", width = 6, height = 9/2)
# p_full_length + theme(legend.position = "top")
# dev.off()
```


## p matched mappable read length
```{r}
rlData <- readRDS(paste0(wkdir,"readLengthData_May6.rds"))
rlData[grepl("PacBio",runname)&grepl("sequin",runname), runname := gsub("v1.0","V2",runname)]
rlData[, old_runname := runname]
temp <- samplesRC_combined[,.(old_runname, runname)]
setnames(temp, "runname", "new_name")
rlData <- temp[rlData, on = "old_runname"]
rlData[!is.na(new_name), runname := new_name]
rlData[, `:=`(old_runname = NULL, new_name = NULL)]
rlData_mat <- samplesRC_combined[rlData, on = "runname"]

rlData_mat[Platform == "Illumina", demultiplexed := TRUE]
rlData_mat[, special_prep:=ifelse(grepl("allSpik",runname),4,ifelse(demultiplexed,3,ifelse(Platform == "PromethION",2,1)))]
rlData_mat[is.na(special_prep), special_prep := 1]
library(ggpubr)
rlData_mat[, nc := all(c("cDNA","directcDNA","directRNA") %in% protocol_type), by = cellLineRep ]
rldata_matched <- unique(rlData_mat[!(protocol_type %in% c("PacBio","Illumina"))&nc&(!is.na(cellLineRep)), list(aveLen = mean(aveLen)), by = list(cellLineRep, protocol_type, protocol_type_factor, cellLine)])
p_mapped_read_length_matched <- ggplot(rldata_matched, 
    aes(x = protocol_type_factor, y = aveLen))+ #/1000
       
    geom_boxplot(col = "black",width = 0.5)+
    #geom_hline(yintercept = c(0.8, 0.9), linetype = "dashed", col = "grey")+
    geom_line(aes(group = cellLineRep), col = "grey", size = 0.4)+
    geom_point(aes(col = cellLine), alpha = 0.5)+
    stat_compare_means(paired = TRUE, method = "t.test", 
                       comparisons = list(c("cDNA","directcDNA"),c("directcDNA","directRNA")))+
     # scale_fill_manual(values = protocolCol[1:3],
     #                  name = "Protocol",
     #                  limits = protocolVec[1:3],
     #                  labels = protocolLabel[1:3])+
    # scale_color_manual(values = protocolCol[1:3],
    #                   name = "Protocol",
    #                   limits = protocolVec[1:3],
    #                   labels = protocolLabel[1:3])+
    scale_color_brewer(type = "qual", palette= 3)+
    xlab("Protocols")+
    ylab("Mean read length(Mappable)")+
    theme_classic()
p_mapped_read_length_matched
```


## q matched read length for one cellline
```{r}
rlData <- readRDS("readLengthData_replicate_matched.rds")
rlData <- unique(rlData[,list(meanLen=max(meanLen,na.rm = TRUE)), by =  list(cellLinerep, protocol_type,tx_name)])
rlData_wide <- dcast(rlData, cellLinerep + tx_name ~ protocol_type, value.var = "meanLen")
rlData_long <- melt(rlData_wide, id.vars = c("cellLinerep","tx_name","directRNA"), measure.vars = c("cDNA","directcDNA"))


txLengths <- general_list$txLengths
rlData_long <- data.table(txLengths)[rlData_long, on = "tx_name"]

library(ggpubr)
Mode <- function(x) {
  ux <- unique(na.omit(x))
  ux[which.max(tabulate(match(x, ux)))]
}
sum_stats <- unique(rlData_long[, list(modeValue = Mode((value-directRNA)/tx_len),
                                meanValue = mean((value-directRNA)/tx_len,na.rm = TRUE)), by = list(cellLinerep, variable)])
p_coverage_difference <- ggplot(rlData_long[grepl("HEYA8", cellLinerep)&(variable == "cDNA")], aes(x = -(value-directRNA)/tx_len*100, fill = cellLinerep))+
    #geom_vline(xintercept = 0, col = "grey", linetype = "dashed")+
    #geom_histogram(position = "identity",bins = 100, alpha = 0.5)+
    geom_density(alpha = 0.5, col = "white")+
    xlab("Transcript coverage reduction percentage (%) \n PCR cDNA comparing to direct RNA")+
    ylab("Density")+
    facet_wrap(~cellLinerep, nrow = 1)+
    geom_vline(data = sum_stats[grepl("HEYA8", cellLinerep)&(variable == "cDNA")], aes(xintercept = modeValue*100), col = "grey", linetype = "dashed")+
    geom_text(data = sum_stats[grepl("HEYA8", cellLinerep)&(variable == "cDNA")], aes(x = modeValue*100, y = 0.03, label = paste0("Mode: ", modeValue)))+
     geom_vline(data = sum_stats[grepl("HEYA8", cellLinerep)&(variable == "cDNA")], aes(xintercept = -meanValue*100), col = "grey", linetype = "dashed")+
    geom_text(data = sum_stats[grepl("HEYA8", cellLinerep)&(variable == "cDNA")], aes(x = -meanValue*100+5, y = 0.02, label = paste0("Mean: ", -round(meanValue*100))))+
    #ylab("Frequency")+
    #xlim(-1,1)+
    theme_classic()
```

## r read length scatter plot 
```{r}
p_coverage_scatterplot <- ggplot(rlData_long[grepl("HEYA8", cellLinerep)&(variable == "cDNA")], aes(x =  directRNA, y = value))+
    # geom_vline(xintercept = 0, col = "grey", linetype = "dashed")+
    # geom_histogram(position = "identity",bins = 100, alpha = 0.5)+
    #geom_point(alpha = 0.5, col = "lightblue")+
    geom_abline(intercept = 0, slope = 1)+
    geom_hex(binwidth = c(0.005, 0.005), size = 0.5, alpha = 0.1, color = "steelblue")+
    geom_density_2d()+
   # scale_fill_gradient2()+
    #geom_density(alpha = 0.5, col = "white")+
    ylab("Read length (PCR cDNA)")+
    xlab("Read length (directRNA)")+
    facet_wrap(~cellLinerep)+
    scale_x_log10()+
    scale_y_log10()+
    #ylab("Frequency")+
    #xlim(-1,1)+
    theme_classic()
```

## supplementary plot
```{r}

pdf("supp_figure_2_draft_17oct2023.pdf", width = 14, height = 20)
# ggarrange(p_expected_data_cost+theme(legend.position="none"), p_expected_throughput+theme(legend.position="none"),
#           p_total_bases+theme(legend.position="none"), p_error_rate_log10+theme(legend.position="none"), 
#           # ggarrange(
#           #           , nrow = 2, align = "hv"),
#           p_coverage_5to3_spikein +theme(legend.position="none"),
#           p_read_coverage_spikein +theme(legend.position="none"),
#           p_read_coverage+theme(legend.position="none"), 
#           p_full_length_filtered+theme(legend.position="none"), 
#           p_transcript_diversity_boxplot+theme(legend.position="none"),
#          p_transcript_diversity_by_geneLength+theme(legend.position="none"),
#          p_transcript_diversity_boxplot_by_geneLength+theme(legend.position="none"),
#           p_cdna_dcdna+theme(legend.position="none"),
#           p_mapped_read_length_matched+theme(legend.position="none"),
#          
#          p_coverage_difference +theme(legend.position="none"),
#          p_coverage_scatterplot +theme(legend.position="none"),
#          as_ggplot(get_legend(p_expected_data_cost)),
#           as_ggplot(get_legend(p_total_bases)),
#          as_ggplot(get_legend(p_read_coverage)),
#          as_ggplot(get_legend(p_mapped_read_length_matched)),
#           labels = c("a","b","c","d",
#                      "e","f","g","h",
#                      "i","j","k","l",
#                      "","","",""),
#           common.legend = TRUE,
#           ncol = 4, nrow = 5, align = "hv")

ggdraw() +
     draw_plot(p_expected_data_cost+theme(legend.position="none"), 0,16/21,1/3,5/21)+
     draw_plot(p_expected_throughput+theme(legend.position="none"),1/3,16/21, 1/3,5/21)+
     draw_plot(p_total_bases+theme(legend.position="none"), 2/3,16/21,1/3,5/21)+
    draw_plot(p_error_rate_log10+theme(legend.position="none"), 0,11/21,1/3,5/21)+
     draw_plot(p_read_coverage+theme(legend.position="none"), 1/3,11/21,1/3,5/21)+
    draw_plot(p_full_length_filtered+theme(legend.position="none"), 2/3,11/21,1/3,5/21)+
    draw_plot(p_transcript_diversity_boxplot+theme(legend.position="none"), 0,6/21,1/3,5/21)+
    draw_plot(p_transcript_diversity_by_geneLength+theme(legend.position="none"), 1/3,6/21,1/3,5/21)+
    draw_plot(p_transcript_diversity_boxplot_by_geneLength+theme(legend.position="none"),2/3,6/21,1/3,5/21)+
    draw_plot(p_mapped_read_length_matched+theme(legend.position="none"),0,1/21,1/4,5/26)+
    draw_plot(p_coverage_difference +theme(legend.position="none"), 1/4,7/42,3/4,5/42)+
    draw_plot( p_coverage_scatterplot +theme(legend.position="none"),1/4,1/21,3/4,5/42)+
    draw_plot(as_ggplot(get_legend(p_expected_data_cost)), 0,0,1/5,1/21)+
    draw_plot( as_ggplot(get_legend(p_expected_throughput)), 1/5,0,1/5,1/21)+
    draw_plot( as_ggplot(get_legend(p_total_bases)), 2/5,0,1/5,1/21)+
    draw_plot(as_ggplot(get_legend(p_read_coverage)), 3/5,0,1/5,1/21)+
    draw_plot(as_ggplot(get_legend(p_mapped_read_length_matched)),4/5,0,1/5,1/21)

dev.off()
```



```{r}
#
pdf("supp_figure_3_draft_170ct2023.pdf", width = 7, height = 5)
ggdraw() +
        draw_plot(p_sequencing_depth_spikein+theme(legend.position="none"),0,4/7, 1/3,3/7)+
    draw_plot(p_mapped_read_length_spikein+theme(legend.position="none"),1/3,4/7, 1/3,3/7)+
draw_plot(p_total_bases_spikein+theme(legend.position="none"), 2/3,4/7,1/3,3/7)+
draw_plot(p_error_rate_log10_spikein+theme(legend.position="none"), 0,1/7,1/3,3/7)+
    draw_plot(p_coverage_5to3_spikein +theme(legend.position="none"),1/3,1/7,1/3,3/7)+
    draw_plot( p_read_coverage_spikein +theme(legend.position="none"), 2/3,1/7,1/3,3/7)+
    
      draw_plot( as_ggplot(get_legend(p_sequencing_depth_spikein)), 0,0,1/2,1/7)+
     draw_plot(as_ggplot(get_legend(p_coverage_5to3_spikein)), 1/2,0,1/2,1/7)
   
dev.off()
```


# Supplementary Table 2
```{r}
dat <- scoreList[[1]][[2]]
write.table(dat[which(setBiasCandidates&abs(score.cdna.drna) > 3.5)][,
        .(geneID, meanCount.cdna,meanCount.drna, score.cdna.drna, score.cdna.drna.p.adjust)],
        file = "Supp_table2.txt",
        col.names = TRUE, row.names = FALSE, sep = ",", quote = FALSE)
```



## read-length matched sample version
```{r}
rlData <- readRDS(paste0(wkdir,"readLengthData_May6.rds"))
rlData[grepl("PacBio",runname)&grepl("sequin",runname), runname := gsub("v1.0","V2",runname)]
rlData[, old_runname := runname]
temp <- samplesRC_combined[,.(old_runname, runname)]
setnames(temp, "runname", "new_name")
rlData <- temp[rlData, on = "old_runname"]
rlData[!is.na(new_name), runname := new_name]
rlData[, `:=`(old_runname = NULL, new_name = NULL)]
rlData_mat <- samplesRC_combined[rlData, on = "runname"]


rlData_mat[, special_prep:=ifelse(grepl("allSpik",runname),4,ifelse(Platform == "PromethION",2,ifelse(demultiplexed,3,
                         1)))]
rlData_mat[is.na(special_prep), special_prep := 1]

```

```{r}
library(ggpubr)
rlData_mat[, nc := all(c("cDNA","directcDNA","directRNA") %in% protocol_type), by = cellLineRep ]
rldata_matched <- unique(rlData_mat[!(protocol_type %in% c("PacBio","Illumina"))&nc&(!is.na(cellLineRep)), list(aveLen = mean(aveLen)), by = list(cellLineRep, protocol_type, protocol_type_factor, cellLine)])
p_mapped_read_length_matched <- ggplot(rldata_matched, 
    aes(x = protocol_type_factor, y = aveLen))+ #/1000
       
    geom_boxplot(col = "black",width = 0.5)+
    #geom_hline(yintercept = c(0.8, 0.9), linetype = "dashed", col = "grey")+
    geom_line(aes(group = cellLineRep), col = "grey", size = 0.4)+
    geom_point(aes(col = cellLine), alpha = 0.5)+
    stat_compare_means(paired = TRUE, method = "t.test", 
                       comparisons = list(c("cDNA","directcDNA"),c("directcDNA","directRNA")))+
     # scale_fill_manual(values = protocolCol[1:3],
     #                  name = "Protocol",
     #                  limits = protocolVec[1:3],
     #                  labels = protocolLabel[1:3])+
    # scale_color_manual(values = protocolCol[1:3],
    #                   name = "Protocol",
    #                   limits = protocolVec[1:3],
    #                   labels = protocolLabel[1:3])+
    scale_color_brewer(type = "qual", palette= 3)+
    xlab("Protocols")+
    ylab("Average read length(Mappable)")+
    theme_classic()
p_mapped_read_length_matched
pdf("read_length_boxplot_matched_replicates.pdf", width = 4, height = 4)
p_mapped_read_length_matched + theme(legend.position = "top")
dev.off()
```

# reviewer response figure

## r3.1 expected cost and throughput 
```{r}
library(ggpubr)
pdf("r3.1_25Oct2023.pdf", width = 7, height = 3.5)
ggarrange(p_expected_data_cost + theme(legend.position = "top"), p_expected_throughput + theme(legend.position = "top"), nrow = 1, ncol =2, align = "hv")
dev.off()
```



## r3.14 coverage bias
```{r}
pdf("r3.14.pdf", width = 10, height = 4.5)
ggdraw() +
     draw_plot(p_mapped_read_length_matched+theme(legend.position="none"), 0,1/4,1.5/5,3/4)+
     draw_plot(p_coverage_difference+theme(legend.position="none"),1.5/5,5/8, 2/5,3/8)+
    draw_plot(p_coverage_scatterplot+theme(legend.position="none"),1.5/5,1/4, 2/5,3/8)+
       draw_plot(p_cdna_drna+theme(legend.position="none"), 3.5/5,1/4,1.5/5,3/4)+
    draw_plot(as_ggplot(get_legend(p_mapped_read_length_matched)), 0,0,1/2,1/4)
dev.off()
```

# Supplementary Text Fig. 1
full-length filtering distance analysis
full-length filtered 
full-length analysis for public dataset from literature papers
```{r}
rin_data_se_files <- dir("/mnt/projects/SGNExManuscript/output_guppy6.4.2/RIN_data/se/", pattern = ".rds", full.names = TRUE)
# check for short read 
type_dist_rin_data <- lapply(rin_data_se_files, function(b){
  print(b)
  rname <- basename(b)
  rin_se <- readRDS(b)
  reads <- data.table(as.data.frame(metadata(rin_se)$distTables[[1]]))
  reads[, multi := .N, by = readClassId]
  reads[, type := ifelse(multi==1&all(equal), "FIM",
                  ifelse(multi==1&(!all(equal)), "SIM",
                  ifelse(multi>1&(all(equal)),"MFIM",
                  ifelse(multi>1&(!all(equal))&(any(equal)),"FSIM",
                  ifelse(multi>1&(all(!equal)),"MSIM",NA))))), by = readClassId]
  reads <- unique(reads[,.(readClassId, readCount, type)], by = NULL)
  
  reads_type_sum <- unique(reads[, list(read_count = sum(readCount)), by = type], by = NULL)
  reads_type_sum[, nTotal := sum(read_count)]
  reads_type_sum[, runname := basename(b)]
  return(reads_type_sum)
})
saveRDS(type_dist_rin_data, file = "type_dist_rin_se.rds")
```


```{r}
rin_data_se_files <- dir("RIN_data/se/", pattern = ".rds", full.names = TRUE)
# check for short read 
type_dist_rin_data_spliced <- lapply(rin_data_se_files, function(b){
  print(b)
  rname <- basename(b)
  rin_se <- readRDS(b)
  reads <- data.table(as.data.frame(metadata(rin_se)$distTables[[1]]))
  reads <- reads[which(!grepl("unspliced", readClassId))]
  reads <- reads[annotationTxId %in% unique(ensemblAnnotations.transcripts[gene_biotype == "protein_coding"]$tx_name)]
  reads[, multi := .N, by = readClassId]
  reads[, type := ifelse(multi==1&all(equal), "FIM",
                  ifelse(multi==1&(!all(equal)), "SIM",
                  ifelse(multi>1&(all(equal)),"MFIM",
                  ifelse(multi>1&(!all(equal))&(any(equal)),"FSIM",
                  ifelse(multi>1&(all(!equal)),"MSIM",NA))))), by = readClassId]
  reads <- unique(reads[,.(readClassId, readCount, type)], by = NULL)
  
  reads_type_sum <- unique(reads[, list(read_count = sum(readCount)), by = type], by = NULL)
  reads_type_sum[, nTotal := sum(read_count)]
  reads_type_sum[, runname := basename(b)]
  return(reads_type_sum)
})
saveRDS(type_dist_rin_data_spliced, file = "type_dist_rin_se_spliced_reads_only_protein_coding.rds")
```



```{r}
rin_match_data <- data.table(runname = c("ERR9834002","ERR9829522","ERR9834006","ERR9834011","ERR9834013"),
                             rin_value = c(9.8, 9.6,9.3,8.4,7.7))
## rin data plot 
type_dist_rin_data <- readRDS("type_dist_rin_se.rds")#_protein_coding
type_dist_rin_data_spliced <- readRDS("type_dist_rin_se_spliced_reads_only_protein_coding.rds")
type_dist_rin_data <- do.call("rbind",type_dist_rin_data)
type_dist_rin_data_spliced <- do.call("rbind",type_dist_rin_data_spliced)
type_dist_rin_data[, spliced_type := "All reads"]
type_dist_rin_data_spliced[, spliced_type := "Spliced reads from \n protein coding genes"]
type_dist <- rbindlist(list(type_dist_rin_data,type_dist_rin_data_spliced), fill = TRUE)
type_dist[, runname := gsub(".rds","",runname)]
type_dist <- rin_match_data[type_dist, on = "runname"]
sum_stats <- unique(type_dist[, list(full_length = sum(.SD[type %in% c("FIM","FSIM","MFIM")]$read_count)), by = list(rin_value,runname, nTotal, spliced_type)], by = NULL)
sum_stats[, percentage := full_length/nTotal]
p_full_length_rin_data <- ggplot(type_dist, aes(x = type, y = read_count/nTotal))+
    #geom_violin(draw_quantiles = NULL, trim = TRUE,adjust = 0.5)+
    geom_point(aes(col = as.factor(rin_value)), size = 4, shape = 16)+
    # scale_fill_manual(values = protocolCol,
    #                   name = "Protocol",
    #                   limits = protocolVec,
    #                   labels = protocolLabel)+
    # scale_color_manual(values = protocolCol,
    #                    name = "Protocol",
    #                    limits = protocolVec[1:5],
    #                    labels = protocolLabel[1:5])+
    scale_color_brewer(type = "seq", direction = 2, palette = 1)+
    ylab("Fraction of reads")+
    scale_x_discrete(limits = c("FIM","SIM","MSIM","FSIM","MFIM"),
                     label = c("Unique\n(Full-length)",
                               "Unique\n(Partial)",
                               "Multi-mapping\n(Partial)",
                               "Multi-mapping\n(Full-length/partial)",
                               "Multi-mapping\n(Full-length)"))+
    facet_wrap(~spliced_type)+
    xlab("Alignment type")+
    theme_classic()
p_full_length_rin_data
pdf("/full_length_rin_data_spliced_reads_protein_coding_genesonly.pdf", width = 8, height = 4)
p_full_length_rin_data + theme(legend.position = "top")
dev.off()
```



# Supplementary Text Fig. 2
adapter presence in cDNA vs dcDNA to see whether the shortened reads are because of incomplete reverse transcription or incomplete sequencing , whether reads are artefacts (pychpper results)

5' 3' distance from reads to annotations
```{r}
sefiles <- dir("RunBambu5Jul/se", pattern = ".rds", full.names = TRUE)
rcfiles <- dir("RunBambu5Jul/rc", pattern = ".rds", full.names = TRUE)
raw_reads_files <- dir("RunBambu5Jul/raw_reads", pattern = ".rds", full.names = TRUE)
rnames <- gsub("_readClassSe_bambuOutput_5Jul_trackReads.rds","",basename(sefiles))

library(bambu)
np <- lapply(rnames, function(rname){
    print(match(rname,rnames))
    temp_se <- readRDS(sefiles[grep(rname,sefiles)])
    raw_reads <- readRDS(raw_reads_files[grep(rname, raw_reads_files)])
    temp_annotations <- rowRanges(temp_se)
    readToTranscriptMap <- data.table(as.data.frame(metadata(temp_se)$readToTranscriptMaps[[1]]))
    readToTranscriptMap[, status := unlist(lapply(compatibleMatches, function(x) ifelse(is.null(x), NA, x)))]
    readToTranscriptMap_1to1 <- readToTranscriptMap[!is.na(status), list(txid = unlist(compatibleMatches)), by= readId]
    readToTranscriptMap_1to1 <- data.table(as.data.frame(rowData(temp_se)))[,.(txid,TXNAME)][readToTranscriptMap_1to1, on = "txid"]
    compare_output <- compareTranscripts(raw_reads[readToTranscriptMap_1to1$readId],
                       temp_annotations[readToTranscriptMap_1to1$TXNAME])
    compare_output <- data.table(compare_output)
    distance_data <- unique(compare_output[, list(tss = min(abs(alternativeTSS)),
                                           tes = min(abs(alternativeTES))), by = queryId])
    saveRDS(distance_data, file = paste0("distanceData/",rname,".rds"))
    
})
```


```{r}
sefiles <- dir("RunBambu5Jul/se", pattern = ".rds", full.names = TRUE)
rnames <- gsub("_readClassSe_bambuOutput_5Jul_trackReads.rds","",basename(sefiles))
distanceData_files <- dir("RunBambu5Jul/distanceData/", full.names = TRUE)
knames <- gsub(".rds","",basename(distanceData_files))
distanceData <- do.call("rbind",lapply(seq_along(distanceData_files), function(k){
    data <- readRDS(distanceData_files[k])
    return(data.table(tss_mean = mean(data$tss, na.rm = TRUE),
                      tss_median = median(data$tss, na.rm = TRUE),
                      tss_sd = sd(data$tss, na.rm = TRUE),
                      tes_mean = mean(data$tes, na.rm = TRUE),
                      tes_median = median(data$tes, na.rm = TRUE),
                      tes_sd = sd(data$tes, na.rm = TRUE),
                      n_total = nrow(data),
                      runname = knames[k]))
}))
saveRDS(distanceData, file = "RunBambu5Jul/summarised_distanceData.rds")


distanceData <- readRDS("RunBambu5Jul/summarised_distanceData.rds")
distanceData[, protocol := gsub("Stranded","", unlist(strsplit(runname,"_"))[3]), by = runname]
p_tss <- ggplot(distanceData[n_total >=400000], aes(x = protocol, y = tss_mean))+
    geom_boxplot()

p_tes <- ggplot(distanceData[n_total >=400000], aes(x = protocol, y = tes_mean))+
    geom_boxplot()

ggarrange(p_tss, p_tes, nrow = 1)



plotdata <- melt(distanceData[n_total >=400000], id.vars = c("runname","n_total","protocol"),
             measure.vars = c("tss_mean","tes_mean"))
labeldata <- unique(plotdata[, list(meanValue = mean(value)), by = list(variable, protocol)])
labeldata[, value := 450]
p_tss_tes <- ggplot(plotdata, aes(x = variable, y = value, fill = protocol, col = protocol))+
    geom_boxplot(alpha = 0.8)+
    geom_point(aes(size = n_total), position = position_dodge(width = 0.75),shape = 16, alpha = 0.5)+
    scale_fill_brewer(type = "qual", palette = 3)+
    scale_color_brewer(type = "qual", palette = 3)+
    scale_size_continuous(trans = "log10")+
      geom_text(data = labeldata, aes(label = round(meanValue,0)), position = position_dodge(width = 0.75), color = "black", size = 4) +
    ylab("Mean distance to annotated start/end")+
    xlab("")+
    theme_classic()+
     theme(legend.position = "top")


pdf(paste0(wkdir, "distance_between_observed_startends_to_annotated_startends.pdf"), width = 6, height = 4)
p_tss_tes
dev.off()
 

> mean(plotdata[protocol == "cDNA"&(variable == "tes_mean")]$value)
[1] 287.28
> mean(plotdata[protocol == "directcDNA"&(variable == "tes_mean")]$value)
[1] 269.4142
> mean(plotdata[protocol == "directRNA"&(variable == "tes_mean")]$value)
[1] 329.5

 sum(plotdata[protocol == "directRNA"&(variable == "tes_mean")]$value*plotdata[protocol == "directRNA"&(variable == "tes_mean")]$n_total)/sum(plotdata[protocol == "directRNA"&(variable == "tes_mean")]$n_total)
 
 
 k <- which(knames %in% "SGNex_A549_directRNA_replicate4_run1")
 tes_A549_directRNA_replicate4_run1_data <- readRDS(distanceData_files[k])
 labeldata <- data.table(value = c(round(mean(na.omit(tes_A549_directRNA_replicate4_run1_data$tes))), median(na.omit(tes_A549_directRNA_replicate4_run1_data$tes))),
                         type = c("Mean", "Median"))
p_tss_tes_histogram_A549_directRNA_replicate4_run1 <- ggplot(tes_A549_directRNA_replicate4_run1_data, aes(x = tes))+
     geom_vline(xintercept = labeldata$value, linetype = "dashed")+
    geom_text(data = labeldata, aes(x = value, y = 120000, label = value))+
     geom_histogram(binwidth = 0.1, col = "lightblue", fill = "lightblue")+scale_x_log10()+theme_classic()
 


pdf(paste0(wkdir, "distance_between_observed_startends_to_annotated_startends_final.pdf"), width = 7, height = 3.5)
ggarrange(p_tss_tes_histogram_A549_directRNA_replicate4_run1,p_tss_tes, nrow = 1, widths = c(0.8,1))
dev.off()
```

# Supplementary Text Fig. 4
fraction of reads cut short: 5', 3' or both ends 
```{r}
seOutput <- readRDS("bambuOutput_May25.rds")
distTablesList <- metadata(seOutput)$distTables
library(BiocFileCache)
bfc1 <- BiocFileCache("/mnt/projects/SGNExManuscript/output_guppy6.4.2/RunBambu22Apr/rc", ask = FALSE)
info <- bfcinfo(bfc1)
annotation_ranges <- rowRanges(seOutput)
data_reads <- lapply(seq_along(distTablesList), function(k){
  print(k)
  reads <- data.table(as.data.frame(distTablesList[[k]]))
  reads[, multi := .N, by = readClassId]
  r_name <- names(distTablesList)[k]
  rc <- readRDS(info[grepl(r_name,info$rname),]$rpath)
  
  rc_ranges <- rowRanges(rc)
  reads_filtered <- reads[!grepl("unidentified",annotationTxId)]
  out <- compareTranscripts(rc_ranges[reads_filtered$readClassId], annotation_ranges[reads_filtered$annotationTxId])
  setnames(out, c("queryId","subjectId"),c("readClassId","annotationTxId"))
  reads_filtered <- data.table(out)[reads_filtered, on = c("readClassId","annotationTxId")]
  thresholdVec <- c(0,1,5,30,50,NA)
  perc_out <- lapply(seq_along(thresholdVec), function(mm){
       threshold_value <- thresholdVec[mm]
      if(mm <=5){
           reads_filtered[, `:=`(V5 = (!alternativeFirstExon&(abs(alternativeTSS)<=threshold_value)),
                            V3 = (!alternativeLastExon&(abs(alternativeTES)<=threshold_value)))]
      }else{
          reads_filtered[, `:=`(V5 = (!alternativeFirstExon),
                            V3 = (!alternativeLastExon))]
      }
      
      reads_filtered[, `:=`(V5and3 = V5&V3)]
      reads_summed <- unique(reads_filtered[,list(prime5 = any(V5&(!grepl("unspliced",readClassId))),
                                                  prime3 = any(V3&(!grepl("unspliced",readClassId))),
                                                  prime5and3 = any(V5and3&(!grepl("unspliced",readClassId))),
                                                  prime5_all = any(V5),
                                                  prime3_all = any(V3),
                                                  prime5and3_all = any(V5and3)),
                                                  by = list(readClassId,readCount)])
      perc_out <- data.table(
          readCount_firstExonCovered = sum(reads_summed$prime5*reads_summed$readCount),
          readCount_lastExonCovered = sum(reads_summed$prime3*reads_summed$readCount),
          readCount_firstlastExonCovered = sum(reads_summed$prime5and3*reads_summed$readCount),
          readCount_firstExonCovered_all = sum(reads_summed$prime5_all*reads_summed$readCount),
          readCount_lastExonCovered_all = sum(reads_summed$prime3_all*reads_summed$readCount),
          readCount_firstlastExonCovered_all = sum(reads_summed$prime5and3_all*reads_summed$readCount),
           nTotal = sum(reads_summed$readCount),
                         nSplicedTotal = sum(reads_summed[which(!grepl("unspliced",readClassId))]$readCount),
                         runname = r_name)
      perc_out[, filter_value := threshold_value]
      return(perc_out)
  })
  perc_out <- do.call("rbind",perc_out)
  perc_out[, protocol:=strsplit(runname, '\\_')[[1]][3], by = runname]
  return(perc_out)
})
saveRDS(data_reads, file = "/mnt/projects/SGNExManuscript/output_guppy6.4.2/first_last_exon_coverage_filtered_version.rds")
```


```{r}
data_reads <- readRDS("/mnt/projects/SGNExManuscript/output_guppy6.4.2/first_last_exon_coverage.rds")
data <- do.call("rbind",data_reads)
data[, protocol := gsub("Stranded","", protocol)]

dataLong <- melt(data, id.vars = c("runname","protocol","nTotal"), measure.vars = colnames(data)[4:6])
ggplot(dataLong, aes(x = protocol, y = value/nTotal, fill = variable))+
     geom_abline(intercept = c(0.8,0.9), slope = 0, col = "grey")+
    geom_boxplot()+
    theme_classic()



## spliced reads only
data_reads <- readRDS("/mnt/projects/SGNExManuscript/output_guppy6.4.2/first_last_exon_coverage_splicedOnly.rds")
data <- do.call("rbind",data_reads)
data[, protocol := gsub("Stranded","", protocol)]

dataLong <- melt(data, id.vars = c("runname","protocol","nSplicedTotal"), measure.vars = colnames(data)[7:9])
ggplot(dataLong, aes(x = protocol, y = value/nSplicedTotal, fill = variable))+
     geom_abline(intercept = c(0.8,0.9), slope = 0, col = "grey")+
    geom_boxplot()+
    theme_classic()

# 1/50bp
data_reads <- readRDS("/mnt/projects/SGNExManuscript/output_guppy6.4.2/first_last_exon_coverage_splicedOnly_1bp_50bp.rds")
data <- do.call("rbind",data_reads)
data[, protocol := gsub("Stranded","", protocol)]

dataLong <- melt(data, id.vars = c("runname","protocol","nSplicedTotal"), measure.vars = colnames(data)[4:6])
ggplot(dataLong, aes(x = protocol, y = value/nSplicedTotal, fill = variable))+
     geom_abline(intercept = c(0.8,0.9), slope = 0, col = "grey")+
    geom_boxplot()+
    theme_classic()


dataLong <- melt(data, id.vars = c("runname","protocol","nTotal"), measure.vars = colnames(data)[10:12])
ggplot(dataLong, aes(x = protocol, y = value/nTotal, fill = variable))+
     geom_abline(intercept = c(0.8,0.9), slope = 0, col = "grey")+
    geom_boxplot()+
    theme_classic()


data_reads <- readRDS("/mnt/projects/SGNExManuscript/output_guppy6.4.2/first_last_exon_coverage_filtered_version.rds")
data <- do.call("rbind",data_reads)
data[, protocol := gsub("Stranded","", protocol)]
filter_threshold <- 30


dataLong <- melt(data[is.na(filter_value)], id.vars = c("runname","protocol","nTotal"), measure.vars = colnames(data)[4:6])
dataLong[, variable:=gsub("readCount_|_all|Covered","",variable)]
p_first_last_exon_coverage <- ggplot(dataLong, aes(x = protocol, y = value/nTotal, fill = factor(variable, levels = c("firstExon","lastExon","firstlastExon"))))+
    geom_hline(yintercept = c(0.8,0.95), col = "grey", linetype = "dashed")+
    geom_boxplot(outlier.size = 1, outlier.color = "grey", outlier.shape = 16)+
    ylim(0.3,1)+
     #ggtitle(paste(filter_threshold,"All reads"))+
    scale_fill_brewer(type = "qual", palette = 1, name = "Start End coverage")+
    ylab("Fraction of reads")+
    theme_classic()+
    theme(legend.position = "top")
pdf(paste0(wkdir, "revisionFigures_guppy6.4.2/figure2/fraction_first_last_exon_coverage.pdf"), width = 6, height = 4)
p_first_last_exon_coverage
dev.off()

dataLong[, perc := value/nTotal]


sum_stats <- unique(dataLong[,list(med_value = median(perc)), by = list(protocol, variable)])
> sum_stats
     protocol      variable med_value
1:  directRNA     firstExon 0.7157390
2: directcDNA     firstExon 0.6439820
3:       cDNA     firstExon 0.8037574
4:  directRNA      lastExon 0.9084847
5: directcDNA      lastExon 0.8110492
6:       cDNA      lastExon 0.8637533
7:  directRNA firstlastExon 0.6325483
8: directcDNA firstlastExon 0.4696291
9:       cDNA firstlastExon 0.6700289


pList <- lapply(unique(data$filter_value)[1:5], function(filter_threshold){
    dataLong <- melt(data[filter_value == filter_threshold], id.vars = c("runname","protocol","nSplicedTotal"), measure.vars = colnames(data)[1:3])
p1 <- ggplot(dataLong, aes(x = protocol, y = value/nSplicedTotal, fill = variable))+
     geom_abline(intercept = c(0.8,0.9), slope = 0, col = "grey")+
    geom_boxplot()+
    ggtitle(paste(filter_threshold,"Spliced reads"))+
    theme_classic()


dataLong <- melt(data[filter_value == filter_threshold], id.vars = c("runname","protocol","nTotal"), measure.vars = colnames(data)[4:6])
p2 <- ggplot(dataLong, aes(x = protocol, y = value/nTotal, fill = variable))+
     geom_abline(intercept = c(0.8,0.9), slope = 0, col = "grey")+
    geom_boxplot()+
     ggtitle(paste(filter_threshold,"All reads"))+
    theme_classic()

p <- ggarrange(p1+theme(legend.position = "none"),p2+theme(legend.position = "none"), nrow = 1)
return(p)
})


ggarrange(plotlist = pList, ncol = 1, nrow = 5)


```




